---
title: 'Accessibility'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "This vignette shows how to calculate and visualize accessibility in R using the `r5r` package."
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Accessibility} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(tolower(Sys.getenv("NOT_CRAN")), "true"),
  out.width = "100%"
)
```

# 1. Introduction 

Accessibility indicators measure how easily opportunities, such as jobs, can be reached from a given location [@levinson2020manual]. This vignette calculates and maps accessibility with the [`r5r` package](https://ipea.github.io/r5r/index.html), using the Porto Alegre (Brazil) sample data included in `r5r`, in two ways:

- **quick:** `r5r::accessibility()`;
- **flexible:** an `r5r` travel time matrix passed to the [{accessibility} package](https://ipea.github.io/accessibility/), which offers a wider range of metrics. 


# 2. Build routable transport network with `build_network()`

### Increase Java memory and load libraries

Set Java memory before loading the packages ([why](https://ipea.github.io/r5r/articles/r5r.html#usage)).

```{r, message = FALSE}
options(java.parameters = "-Xmx2G")

library(r5r)
library(accessibility)
library(sf)
library(data.table)
library(ggplot2)
library(interp)
library(h3jsr)
library(dplyr)
```
`build_network()` takes the directory holding the OpenStreetMap and GTFS data:

```{r, message = FALSE}
# system.file returns the directory with example data inside the r5r package
# set data path to directory containing your own data if not running this example
data_path <- system.file("extdata/poa", package = "r5r")

r5r_network <- build_network(data_path)
```


# 3. Accessibility: quick and easy approach

One of the simplest accessibility metrics is the cumulative-opportunity metric: the number of opportunities reachable within a travel time cutoff (`decay_function = "step"`).

Here we count the schools (`points$schools`) and public healthcare facilities (`points$healthcare`) reachable by public transport in less than 20 minutes, using the median travel time of departures every minute over a 60-minute window (2pm to 3pm). `accessibility()` handles several opportunity types in one call, which is much more efficient than computing a travel time matrix and aggregating it yourself.


```{r, message = FALSE, warning = FALSE}
# read all points in the city
points <- fread(file.path(data_path, "poa_hexgrid.csv"))

# routing inputs
mode <- c("WALK", "TRANSIT")
max_walk_time <- 30      # in minutes
travel_time_cutoff <- 20 # in minutes
time_window <- 60        # in minutes
departure_datetime <- as.POSIXct("13-05-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")

# calculate accessibility
access1 <- r5r::accessibility(
  r5r_network,
  origins = points,
  destinations = points,
  mode = mode,
  opportunities_colnames = c("schools", "healthcare"),
  decay_function = "step",
  cutoffs = travel_time_cutoff,
  departure_datetime = departure_datetime,
  max_walk_time = max_walk_time,
  time_window = time_window,
  progress = FALSE
  )

head(access1)
```


`r5r::accessibility()` also computes gravity-based metrics: set `decay_function` to `"exponential"`, `"fixed_exponential"`, `"linear"` or `"logistic"`. Metrics not implemented in R5, such as floating catchment area, travel cost to the closest N opportunities or time-interval cumulative opportunities, are available in the `{accessibility}` package.


# 4. Accessibility: flexible approach

The `{accessibility}` package takes a [travel time matrix](https://ipea.github.io/r5r/articles/travel_time_matrix.html), which we compute with `r5r`:

```{r, message = FALSE}
# calculate travel time matrix
ttm <- r5r::travel_time_matrix(
  r5r_network,
  origins = points,
  destinations = points,
  mode = mode,
  departure_datetime = departure_datetime,
  max_walk_time = max_walk_time,
  time_window = time_window,
  progress = FALSE
  )

head(ttm)
```
`accessibility::cumulative_cutoff()` computes the same cumulative metric from the travel time matrix and land use data. It also counts trips that take exactly the cutoff time, while `r5r::accessibility()` counts only trips strictly shorter than it, so the two estimates can differ slightly:

```{r, message = FALSE}
# calculate accessibility
access_edu <- accessibility::cumulative_cutoff(
  travel_matrix = ttm, 
  land_use_data = points,
  opportunity = 'schools',
  travel_cost = 'travel_time_p50',
  cutoff = 20
  )

access_health <- accessibility::cumulative_cutoff(
  travel_matrix = ttm, 
  land_use_data = points,
  opportunity = 'healthcare',
  travel_cost = 'travel_time_p50',
  cutoff = 20
  )

head(access_edu)
head(access_health)
```


# 5. Map Accessibility

Two ways to map the estimates:

## 5.1 Choropleth maps

Each point is the centroid of a fine-resolution H3 hexagon, so we retrieve the hexagon polygons and join the accessibility estimates to them.

```{r, message = FALSE, out.width='100%'}
# retrieve polygons of H3 spatial grid
grid <- h3jsr::cell_to_polygon(points$id, simple = FALSE)

# merge accessibility estimates
access_sf <- left_join(grid, access1, by = c('h3_address'='id'))

# plot
ggplot() +
  geom_sf(data = access_sf, aes(fill = accessibility), color= NA) +
  scale_fill_viridis_c(direction = -1, option = 'B') +
  labs(fill = "Number of\nfacilities within\n20 minutes") +
  theme_minimal() +
  theme(axis.title = element_blank()) +
  facet_wrap(~opportunity) +
  theme_void()

```


## 5.2 Spatial interpolation

Alternatively, interpolate the point estimates to get a smoother surface:

```{r, message = FALSE, out.width='100%'}
# interpolate estimates to get spatially smooth result
access_schools <- access1 %>% 
  filter(opportunity == "schools") %>%
  inner_join(points, by='id') %>%
  with(interp::interp(lon, lat, accessibility)) %>%
  with(cbind(acc=as.vector(z),  # Column-major order
             x=rep(x, times=length(y)),
             y=rep(y, each=length(x)))) %>% as.data.frame() %>% na.omit() %>%
  mutate(opportunity = "schools")

access_health <- access1 %>% 
  filter(opportunity == "healthcare") %>%
  inner_join(points, by='id') %>%
  with(interp::interp(lon, lat, accessibility)) %>%
  with(cbind(acc=as.vector(z),  # Column-major order
             x=rep(x, times=length(y)),
             y=rep(y, each=length(x)))) %>% as.data.frame() %>% na.omit() %>%
  mutate(opportunity = "healthcare")

access.interp <- rbind(access_schools, access_health)

# find results' bounding box to crop the map
bb_x <- c(min(access.interp$x), max(access.interp$x))
bb_y <- c(min(access.interp$y), max(access.interp$y))

# extract OSM network, to plot over map
street_net <- street_network_to_sf(r5r_network)

# plot
ggplot(na.omit(access.interp)) +
  geom_sf(data = street_net$edges, color = "gray55", size=0.01, alpha = 0.7) +
  geom_contour_filled(aes(x=x, y=y, z=acc), alpha=.7) +
  scale_fill_viridis_d(direction = -1, option = 'B') +
  scale_x_continuous(expand=c(0,0)) +
  scale_y_continuous(expand=c(0,0)) +
  coord_sf(xlim = bb_x, ylim = bb_y, datum = NA) + 
  labs(fill = "Number of\nfacilities within\n20 minutes") +
  theme_void() +
  facet_wrap(~opportunity)
```


### Cleaning up after usage

Stop the network and run Java's garbage collector to free the memory it used:

```{r, message = FALSE}
r5r::stop_r5(r5r_network)
rJava::.jgc(R.gc = TRUE)
```

If you have any suggestions or want to report an error, please visit [the package GitHub page](https://github.com/ipea/r5r).

## References
