---
title: 'Building a custom network with modified OSM car speeds'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "This vignette shows how to build a custom transport network with modified OSM car speeds, which can be used to simulate different scenarios of traffic congestion and road closures."
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Modify OSM car speeds} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(tolower(Sys.getenv("NOT_CRAN")), "true"),
  out.width = "100%"
)

```

# 1. Introduction 

By default, routing by car in `R5` considers that vehicles travel at the legal speed limit in each OSM road edge. This is commonly referred to as a "free flow scenario", without congestion. However, the average speed of car trips is different (usually lower) than the legal speed limit in most real case scenarios due to traffic conditions and driving behavior.

This vignette shows how you can set custom speeds for cars by making simple changes to OSM data with the convenient `build_custom_network()` function. We'll show how OSM speeds can be changed using two different strategies: (1) by applying different speed factors by road hierarchy, and (2) by changing the same speed factor to all roads. A third scenario would be to change the speed of only the roads within determined areas/polygons, but we'll leave this for a future occasion.


## Basic usage

The `build_custom_network()` has a relatively similar behavior to the `r5r::build_network()` function. It builds a transport network that can be used in the routing and accessibility function in `r5r`. The two key differences here are:

1. The user needs to pass a directory path where the new transport network is going to be saved. By default, `build_custom_network()` builds the new network in a temporary directory, but users can point to a permanent directory using the `output_dir` parameter.
2. The user needs to point to pass to the `new_carspeeds` parameter a `data.frame` that indicates the speed factor for each OSM edge id.

In this first example we'll be using the a sample data set for the city of Porto Alegre (Brazil) included in `r5r`. This sample data path includes the file `"poa_osm_congestion.csv"`, which should have two columns: `"osm_id"` and `"max_speed"`. Like this:

```{r, message=FALSE}
# increase Java memory
options(java.parameters = "-Xmx2G")

# load libraries
library(r5r)
library(dplyr)
library(data.table)
library(ggplot2)

# data path where the .pbf file is located
data_path <- system.file("extdata/poa", package = "r5r")

# read data.frame with speed factors
speed_factors <- read.csv(
  system.file("extdata/poa/poa_osm_congestion.csv", package = "r5r")
  )

head(speed_factors)

```
In this example, the values of the `"max_speed"` column in the `speed_factors` object are all set to `0.5`. Since we also set the parameter `percentage_mode = TRUE`, this means that the driving speed of those OSM edges listed in the `data.frame` will be at 50% of the original speed in the OSM data. Mind you that the values in the `"max_speed"` column also accept absolute values in Km/h, in which case you must use `percentage_mode = FALSE`.

obs. by default, the speeds of all the OSM edges not listed in the `data.frame` are kept unchanged. You can change the default value, as we'll see in the second example below.

```{r, message = FALSE}
r5r_network_congestion <- r5r::build_custom_network(
  data_path = data_path,
  new_carspeeds = speed_factors,
  percentage_mode = TRUE,
  verbose = FALSE
  )

```

And that's it ! You can now use the new network `r5r_network_congestion` in any of `r5r`'s routing and accessibility functions, like this.

```{r, eval=FALSE}
# load origin/destination points
points <- data.table::fread(file.path(data_path, "poa_points_of_interest.csv"))

ttm_congestion <- r5r::travel_time_matrix(
  r5r_network_congestion,
  origins = points,
  destinations = points,
  mode = 'car',
  departure_datetime = Sys.time(),
  max_trip_duration = 60
)
```
Obs. Mind you however that, even though we have set the speed factors to `0.5`, travel times might not become twice as long. This is because of how travel times by car are also affected by intersections, and how changes in the road speeds might also affect the route and hence the trip distance itself.  

Now let's dive into more realistic examples.

# 2. Setting different congestion levels by road hierarchy

In this example, we'll set different speed factor for roads of different hierarchy levels. We can assume for example that congestion levels tend to be more intense in roads of higher hierarchy. We can do this in two simple steps.

First we need to do read the OSM data from our `.pbf` file, and to filter the OSM edges with the road types we want.


```{r, message=FALSE}
# path to OSM pbf
pbf_path <- paste0(data_path, "/poa_osm.pbf")
  
# read layer of lines from pbf
roads <- sf::st_read(
  pbf_path, 
  layer = 'lines', 
  quiet = TRUE
  )

# Filter only road types of interest
rt <- c("motorway", "primary", "secondary", "tertiary") 

roads <- roads |>
  select(osm_id, highway) |>
  filter(highway %in% rt)

head(roads)
```

Here's how the road network looks like.

```{r}
# map
plot(roads["highway"])

```

Now we only need to add a new column `"max_speed"` with values conditioned on the road type, and make sure the `osm_id` is of class `numeric`. The `data.frame` looks like this:

```{r}
new_speeds <- roads |>
  mutate( 
    osm_id = as.numeric(osm_id),
    max_speed = case_when(
      highway == "motorway"  ~ 0.75,
      highway == "primary"   ~ 0.8,
      highway == "secondary" ~ 0.85,
      highway == "tertiary"  ~ 0.9)) |>
  sf::st_drop_geometry()

head(new_speeds)

```

That's it. Now we can pass the `new_speeds` data frame to `build_custom_network()` along with the path to the directory where the original OSM `.pbf` file is stored and the function will build a new routable transport network with modified OSM car speeds.

```{r, message = FALSE}

r5r_network_new_speeds <- r5r::build_custom_network(
  data_path = data_path,
  new_carspeeds = new_speeds,
  output_path = tempdir(),
  percentage_mode = TRUE,
  verbose = FALSE
  )
```


# 3. Applying the same speed factor to all roads

The `build_custom_network()` function has a parameter `default_speed`, which can be used to set the speed of all the roads not listed in the `new_carspeeds` input. By default, `default_speed = NULL` and the speeds of the unlisted roads are kept unchanged.

In this example, we'll use this parameter to apply the same speed factor to all roads, to simulate as if the speed limits of all roads was changed to 40 Km/h. To do this, we only need to create a `data.frame` with a mock OSM id that that does not exist in our OSM data and to pass `default_speed = 40` with `percentage_mode = FALSE`.


```{r, message = FALSE}
# create df with a road that does not exist.
mock_data <- data.frame(osm_id = 9999, max_speed = 9999)

# create a temp dir where the custom network will be saved
new_temp_dir <- fs::path_temp(tmpdir='./40kmph')
dir.create(new_temp_dir)

# build network with all roads at at 40 Km/h 
r5r_network_20kmph <- r5r::build_custom_network(
  data_path = data_path,
  new_carspeeds = mock_data,
  output_path = new_temp_dir,
  default_speed = 40,
  percentage_mode = FALSE
  )
```

# 4. Congestion level within a polygon

steps TODO:

- create a custom polygon, e.g. a circle over the city center
- determine OSM ids in the polygon using a spatial join operation between the polygon and the `.pbf`
- create a csv file with custom speeds for OSM ids within the polygon

this can be computationally intensive due to the spatial join operation, depending on the size of the road network and the number of polygons. A new function under development will make this much easier / faster.

# Extra tip: 

- **Road closure**: one can simulate a road closure by setting the `"max_speed"` value to `0`. This can be quite handy for studies that try to measure the resilience of transport systems to network disruptions.


### Cleaning up after usage

`r5r` objects are still allocated to any amount of memory previously set after they are done with their calculations. In order to remove an existing `r5r` object and reallocate the memory it had been using, we use the `stop_r5` function followed by a call to Java's garbage collector, as follows:

```{r, message = FALSE}
r5r::stop_r5(r5r_network_congestion)
rJava::.jgc(R.gc = TRUE)
```

```{r, eval = TRUE, include = FALSE, message = FALSE}
# clean cache (CRAN policy)
r5r::r5r_cache(delete_file = 'all')

```

If you have any suggestions or want to report an error, please visit [the package GitHub page](https://github.com/ipeaGIT/r5r).

