---
title: 'Building a custom network with modified OSM car speeds'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "This vignette shows how to build a custom transport network with modified OSM car speeds, which can be used to simulate different scenarios of traffic congestion and road closures."
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Modify OSM car speeds} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(tolower(Sys.getenv("NOT_CRAN")), "true"),
  out.width = "100%"
)

```

# 1. Introduction 
By default, routing by car in `R5` considers that vehicles travel at the legal speed limit in each OSM road edge. This is commonly referred to as a "free flow scenario", without congestion. However, the average speed of car trips is different (usually slower) than the legal speed limit in most real case scenarios due to traffic conditions and driving behavior.

This vignette shows how you can set custom speeds for cars by making simple changes to OSM data with the convenient `build_custom_network()` function. OSM speeds can be changed using two different strategies: (1) by applying different speed factors to each road, or (2) by applying speed factors only to the roads within determined polygons.

The `build_custom_network()` has a relatively similar behavior to the `r5r::build_network()` function. It builds a transport network that can be used in the routing and accessibility function in `r5r`. The two key differences here are:

1. The user needs to pass a directory path where the new transport network is going to be saved. This is done with the `output_path` parameter. In this vignnete, we will simply use the default behavior of the `build_custom_network()` function that builds the new network in a temporary directory.
2. The user needs to pass some info to the `new_carspeeds`, which can be either a `data.frame` that indicates the speed factor for each OSM edge id, OR an `sf data.frame` polygon that indicates the speed factor for all the roads that fall within that polygon.

Let's see how this works with a few examples using the a sample data set for the city of Porto Alegre (Brazil) included in `r5r`.

There are two congestion modes:
- **Edge mode:** The new speed factors must be passed as a `data.frame`
indicating the new max speed of each OSM edge id. The table must contain
columns \code{osm_id} and \code{max_speed}. Values in \code{max_speed} can be
interpreted as percentages of original speeds or as absolute speeds (km/h) by
using `percentage_mode` parameter.
- **Polygon mode:** The new speed factors must be passed through "congestion
region(s)" specified as an `sf data.frame` containing  in each row an
`sf polygon` outlining the "congestion region", \code{scale} and \code{priority}.
Values in \code{scale} can only be interpreted as percentages of original
speeds as `percentage_mode` must be `TRUE`. \code{Priority} is used to apply
a hierarchy in case of overlapping polygons.

# 2. Changing car speeds by OSM edge

## 2.1 Basic usage

In this first example, we will pass the new car speeds using a sample `data.frame` that comes with the package. Mind you that this data frame must have the two columns: `"osm_id"` and `"max_speed"`. Like this:

```{r, message=FALSE}
# increase Java memory
options(java.parameters = "-Xmx2G")

# load libraries
library(r5r)
library(dplyr)
library(data.table)
library(ggplot2)

# data path where the .pbf file is located
data_path <- system.file("extdata/poa", package = "r5r")

# read data.frame with speed factors
speed_factors <- read.csv(
  system.file("extdata/poa/poa_osm_congestion.csv", package = "r5r")
  )

head(speed_factors)

```
In this example, the values of the `"max_speed"` column are all set to `0.5`. Since we also set the parameter `percentage_mode = TRUE`, this means that the driving speed of those OSM edges listed in the `data.frame` will be at 50% of the original speed in the OSM data. Mind you that the values in the `"max_speed"` column also accept absolute values in Km/h, in which case you must use `percentage_mode = FALSE`.

Now to build a new routeable transport network consideering these new driving speeds, you simply need to call the `r5r::build_custom_network()` function:

```{r, message = FALSE}
r5r_network_congestion1 <- r5r::build_custom_network(
  data_path = data_path,
  new_carspeeds = speed_factors,
  percentage_mode = TRUE,
  verbose = FALSE
  )

```

And that's it ! You can now use the new network `r5r_network_congestion1` in any of `r5r`'s routing and accessibility functions, like this.

```{r, eval=FALSE}
# load origin/destination points
points <- data.table::fread(file.path(data_path, "poa_points_of_interest.csv"))

ttm_congestion <- r5r::travel_time_matrix(
  r5r_network_congestion1,
  origins = points,
  destinations = points,
  mode = 'car',
  departure_datetime = Sys.time(),
  max_trip_duration = 30
)
```
Obs. Mind you however that, even though we have set the speed factors to `0.5`, travel times might not become twice as long. This is because of how travel times by car are also affected by intersections, and how changes in the road speeds might also affect the route and hence the trip distance itself.  

Now let's dive into more realistic examples.

## 2.2 Setting different congestion levels by road hierarchy

In this example, we'll set different speed factor for roads of different hierarchy levels. We can assume for example that congestion levels tend to be more intense in roads of higher hierarchy. We can do this in two simple steps.

First we need to do read the OSM data from our `.pbf` file, and to filter the OSM edges with the road types we want.


```{r, message=FALSE}
# path to OSM pbf
pbf_path <- paste0(data_path, "/poa_osm.pbf")
  
# read layer of lines from pbf
roads <- sf::st_read(
  pbf_path, 
  layer = 'lines', 
  quiet = TRUE
  )

# Filter only road types of interest
rt <- c("motorway", "primary", "secondary", "tertiary") 

roads <- roads |>
  select(osm_id, highway) |>
  filter(highway %in% rt)

head(roads)
```

Here's how the road network looks like.

```{r}
# map
plot(roads["highway"])

```

Now we only need to add a new column `"max_speed"` with values conditioned on the road type, and make sure the `osm_id` is of class `numeric`. The `data.frame` looks like this:

```{r}
new_edge_speeds <- roads |>
  mutate( 
    osm_id = as.numeric(osm_id),
    max_speed = case_when(
      highway == "motorway"  ~ 0.75,
      highway == "primary"   ~ 0.8,
      highway == "secondary" ~ 0.85,
      highway == "tertiary"  ~ 0.9)) |>
  sf::st_drop_geometry()

head(new_edge_speeds)

```

That's it. Now we can pass the `new_speeds` data frame to `build_custom_network()` along with the path to the directory where the original OSM `.pbf` file is stored and the function will build a new routable transport network with modified OSM car speeds.

```{r, message = FALSE}

r5r_network_congestion2 <- r5r::build_custom_network(
  data_path = data_path,
  new_carspeeds = new_edge_speeds,
  output_path = tempdir(),
  percentage_mode = TRUE,
  verbose = FALSE
  )
```


## 2.3. Applying the same speed factor to all roads

The `build_custom_network()` function has a parameter `default_speed`, which can be used to set the speed of all the roads not listed in the `new_carspeeds` input. By default, `default_speed = NULL` and the speeds of the unlisted roads are kept unchanged.

In this example, we'll use this parameter to apply the same speed factor to all roads to simulate as if the speed limits of all roads were changed to 40 Km/h. To do this, we only need to create a `data.frame` with a mock OSM id that that does not exist in our OSM data and to pass `default_speed = 40` with `percentage_mode = FALSE`.


```{r, message = FALSE}
# create df with a road that does not exist.
mock_data <- data.frame(osm_id = 9999, max_speed = 9999)

# create a temp dir where the custom network will be saved
new_temp_dir <- fs::path_temp(tmpdir='./40kmph')
dir.create(new_temp_dir)

# build network with all roads at at 40 Km/h 
r5r_network_20kmph <- r5r::build_custom_network(
  data_path = data_path,
  new_carspeeds = mock_data,
  output_path = new_temp_dir,
  default_speed = 40,
  percentage_mode = FALSE
  )
```

### Extra tip: 

- **Road closure**: one can simulate a road closure by setting the `"max_speed"` value to `0`. This can be quite handy for studies that try to measure the resilience of transport systems to network disruptions.


# 3. Changing car speeds with a spatial polygon

If you do not want to set the speed factor for each individual OSM road edge, is to use one or more spatial polygons to set speed the speed factors of all the roads within those polygons. In this example with the sample data from {r5r}, we have two polygons in the city of Porto Alegre. The first one is covers the extended city center, and the second polygon covers a few important roads that connect two major avenues in the city.

```{r, message = FALSE}
# read sf with congestion polygons
congestion_poly <- readRDS(file.path(data_path, "poa_poly_congestion.rds"))

# preview
mapview::mapview(congestion_poly, zcol="scale")
```

Mind you that this `sf data.frame` must have a few mandatory columns:

- `"poly_id"`: a unique id for each polygon
- `"scale"`: the speed scaling factor for each polygon. Notice that this parameter only works with relative speed factors so that `percentage_mode = TRUE` always.
- `"priority"`: a number ranking which polygon should be considered in case of overlapping polygons.

```{r, message = FALSE}
head(congestion_poly)
```
In this example, we are simularing a higher congestion level of the roads in the city center, which would be running at 70% of the legal speed limit, and a slightly better performance for the roads in the second polygon, running at 80% of the speed limit. Finally, we can set `default_speed = 0.95` to simulate that all other roads in the city would be running at 95%.

```{r, message = FALSE}
r5r_network_congested_areas <- r5r::build_custom_network(
  data_path = data_path,
  new_carspeeds = congestion_poly,
  default_speed = 0.95,
  percentage_mode = TRUE
  )
```

And that's it!

### Cleaning up after usage

`r5r` objects are still allocated to any amount of memory previously set after they are done with their calculations. In order to remove an existing `r5r` object and reallocate the memory it had been using, we use the `stop_r5` function followed by a call to Java's garbage collector, as follows:

```{r, message = FALSE}
# stop an specific r5r network
r5r::stop_r5(r5r_network_congestion1)

# or stop all r5r networks at once
r5r::stop_r5()
rJava::.jgc(R.gc = TRUE)
```

```{r, eval = TRUE, include = FALSE, message = FALSE}
# clean cache (CRAN policy)
r5r::r5r_cache(delete_file = 'all')

```

If you have any suggestions or want to report an error, please visit [the package GitHub page](https://github.com/ipeaGIT/r5r).

