---
title: 'r5r: Rapid Realistic Routing with R5 in R'
author: "Rafael H. M. Pereira, Pedro R. Andrade, Joao Bazzo"
output: github_document
abstract: `r5r` is an R package for rapid realistic routing on multimodal 
transport networks (walk, bike, public transport and car) using R5 
<https://github.com/conveyal/r5>. The package allows users to generate detailed
routing analysis or calculate travel time matrices using seamless parallel 
computing on top of the R5 Java machine.
urlcolor: blue
vignette: |
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{r5r}
---
15 August 2020

# Introduction 

**r5r** is an R package for rapid realistic routing on multimodal transport 
networks (walk, bike, public transport and car). It provides a simple and 
friendly interface to R5, the [Rapid Realistic Routing on Real-world and Reimagined networks](https://github.com/conveyal/r5).

# Installation
To use `r5r`, you need to have [Java SE Development Kit 11.0.8](https://www.oracle.com/java/technologies/javase-jdk11-downloads.html) 
installed on your computer. No worries, you don't have to pay for it.

Install the development version from GitHub (soon on CRAN).

```{r, eval = FALSE}
devtools::install_github("ipeaGIT/r5r", subdir = "r-package")
library(r5r)
library(sf)
library(data.table)
library(dplyr)

```

# Demonstration on sample data

## Data
To illustrate functionality, the package includes a small sample data for the 
city of Porto Alegre (Brazil). It includes three files:
- An Open Street Map network in `.pbf` format (*mandatory*)
- A public transport network in `GTFS.zip` format
- The spatial coordinates of the origin destination pairs in a `.csv` file.

```{r}
path <- system.file("extdata", package = "r5r")
list.files(path)

```
The data with origin destination pairs looks like this below. In this example, 
we will be looking at the transport routes between ten random points in this 
data.
```{r}
points <- fread(system.file("extdata/poa_hexgrid.csv", package = "r5r"))
points <- points[ c(sample(1:nrow(points), 10, replace=TRUE)), ]
head(points)

```

## Build routable transport network

The first step is to build the multimodal transport used for routing in R5. This
is done with the `setup_r5` function. It does two things: (1) downloads a compiled JAR file
of R5 and store it locally in the r5r package directory; (2) combines the osm.pbf and gtfs.zip data sets to build a routable network object. The function 

```{r, message = FALSE}
# Indicate the path where OSM and GTFS data are stored
r5_core <- setup_r5(data_path = path)

```

## Routing analysi

**r5** has two core routing functions.

- `detailed_itineraries`
   * Returns a `data.frame sf LINESTRINGs` with one or multiple alternative routes
   between one or multiple origin destination pairs. The data output brings 
   detailed information on transport mode, travel time, walk distance etc for 
   each trip section
 
- `travel_time_matrix`
   * Fast function that returns a simple 'data.frame' with travel time 
   estimates between one or multiple origin destination pairs.
   
### Detailed itineraries

### Fast many to many travel time matrix


```{r, message = FALSE}
# calculate a travel time matrix
ttm <- travel_time_matrix( r5_core = r5_core,
                           origins = points,
                           destinations = points,
                           trip_date = "2019-05-20",
                           departure_time = "14:00:00",
                           direct_modes = c("WALK", "BICYCLE", "CAR"),
                           transit_modes = "BUS",
                           max_street_time = 300L,
                           max_trip_duration = 3600L
                          )
head(ttm)
```



# Final remarks

If you have any suggestions or want to report an error, please visit the GitHub page of the package [here](https://github.com/ipeaGIT/r5r).
