---
title: 'Isochrones'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "This vignette shows how to calculate and visualize isochrones in R using the `r5r` package."
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Isochrones} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(tolower(Sys.getenv("NOT_CRAN")), "true"),
  out.width = "100%"
)
```

# 1. Introduction

An isochrone shows all areas reachable from a place within a maximum travel time. This vignette uses the [`r5r` package](https://ipea.github.io/r5r/index.html) and the Porto Alegre (Brazil) sample data included in it to calculate and map public transport isochrones from the city's central bus station at several travel time thresholds. `r5r::isochrone()` builds both polygon- and line-based isochrones; we cover both.

***Warning:*** to count opportunities (e.g. jobs, schools or hospitals) within each isochrone, we strongly recommend NOT using `isochrone()`. The [Accessibility vignette](https://ipea.github.io/r5r/articles/accessibility.html) shows much more efficient ways.


# 2. Build routable transport network with `build_network()`

### Increase Java memory and load libraries

Set Java memory before loading the packages ([why](https://ipea.github.io/r5r/articles/r5r.html#usage)).

```{r, message = FALSE}
options(java.parameters = "-Xmx2G")

library(r5r)
library(sf)
library(data.table)
library(ggplot2)
```


`build_network()` takes the directory holding the OpenStreetMap and GTFS data:

```{r, message = FALSE}
# system.file returns the directory with example data inside the r5r package
# set data path to directory containing your own data if not running this example
data_path <- system.file("extdata/poa", package = "r5r")

r5r_network <- build_network(data_path)
```

# 3. Calculating and visualizing isochrones 


## 3.1 Polygon-based isochrones 

Polygon-based isochrones are the most common: set `polygon_output = TRUE`. The polygons are built on a regular grid of [Web Mercator pixels](https://docs.conveyal.com/analysis/methodology#spatial-resolution) whose resolution is set by `zoom`. Higher zooms give more detailed isochrones but take longer, and may fail with an error in large networks. The default, 10, uses cells of 153 m at the Equator.

Below, isochrones use the median travel time of departures every minute over a 60-minute window (2pm to 3pm).

```{r, message = FALSE}
# read all points in the city
points <- fread(file.path(data_path, "poa_hexgrid.csv"))

# subset point with the geolocation of the central bus station
central_bus_stn <- points[291,]

# isochrone intervals
time_intervals <- seq(0, 100, 10)

# routing inputs
mode <- c("WALK", "TRANSIT")
max_walk_time <- 30      # in minutes
time_window <- 60        # in minutes
departure_datetime <- as.POSIXct("13-05-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")

# calculate travel time matrix
iso1 <- r5r::isochrone(
  r5r_network,
  origins = central_bus_stn,
  mode = mode,
  polygon_output = TRUE, 
  cutoffs = time_intervals,
  departure_datetime = departure_datetime,
  max_walk_time = max_walk_time,
  time_window = time_window,
  progress = FALSE,
  zoom = 10
  )

```
`isochrone()` works like `travel_time_matrix()`, but with `polygon_output = TRUE` it returns an `sf` data.frame with one `POLYGON`/`MULTIPOLYGON` per origin, cutoff and percentile:

```{r, message = FALSE}
head(iso1)
```

Mapping the isochrones:

```{r, message = FALSE}
# extract OSM network
street_net <- street_network_to_sf(r5r_network)
main_roads <- subset(street_net$edges, street_class %like% 'PRIMARY|SECONDARY')
  
colors <- c('#ffe0a5','#ffcb69','#ffa600','#ff7c43','#f95d6a',
            '#d45087','#a05195','#665191','#2f4b7c','#003f5c')

ggplot() +
  geom_sf(data = iso1, aes(fill=factor(isochrone)), color = NA, alpha = .7) +
  geom_sf(data = main_roads, color = "gray55", size=0.01, alpha = 0.2) +
  geom_point(data = central_bus_stn, aes(x=lon, y=lat, color='Central bus\nstation')) +
  scale_fill_manual(values = rev(colors) ) +
  scale_color_manual(values=c('Central bus\nstation'='black')) +
  labs(fill = "Travel time\n(in minutes)", color='') +
  theme_minimal() +
  theme(axis.title = element_blank())
```
## 3.2 Line-based isochrones 

For line-based isochrones, set `polygon_output = FALSE` (no `zoom` needed). The output is a `LINESTRING` `sf` data.frame.


```{r, message = FALSE}
# calculate travel time matrix
iso2 <- r5r::isochrone(
  r5r_network,
  origins = central_bus_stn,
  mode = mode,
  polygon_output = FALSE, 
  cutoffs = time_intervals,
  departure_datetime = departure_datetime,
  max_walk_time = max_walk_time,
  time_window = time_window,
  progress = FALSE
  )

head(iso2)

```

Mapping the isochrones:

```{r, message = FALSE}
ggplot() +
  geom_sf(data = iso2, aes(color=factor(isochrone)), alpha = .7) +
  scale_color_manual(values = rev(colors) ) +
  geom_point(data = central_bus_stn, aes(x=lon, y=lat), color='black') +
  labs(color = "Travel time\n(in minutes)") +
  theme_minimal() +
  theme(axis.title = element_blank())
```


### Cleaning up after usage

Stop the network and run Java's garbage collector to free the memory it used:

```{r, message = FALSE}
r5r::stop_r5(r5r_network)
rJava::.jgc(R.gc = TRUE)
```

If you have any suggestions or want to report an error, please visit [the package GitHub page](https://github.com/ipea/r5r).
