---
title: 'Analyzing the trade-offs between travel time and monetary cost'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "This vignette shows how to use the `pareto_frontier()` function to examine the trade-offs between travel time and monetary cost in travel time matrices in r5r."
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Analyzing the trade-offs between travel time and monetary cost} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(tolower(Sys.getenv("NOT_CRAN")), "true"),
  out.width = "100%"
)
```

# 1. Introduction

In most cases, transport routing models find either the fastest or the lowest-cost
routes that connect places in a given transport network. Sometimes, though, we 
might want a more sophisticated analysis that considers both the time and monetary 
costs that public transport passengers have to face. The problem here is that simultaneously accounting for both time and monetary costs is a major challenge 
for routing models because of the trade-offs between the objectives of minimizing 
trip duration and cost [(ref 'Getting Charlie off the MTA')](https://www.tandfonline.com/doi/abs/10.1080/13658816.2019.1605075).

To address this problem, `r5r` has a function called `pareto_frontier()`, which calculates the most efficient route possibilities between origin destination pairs considering multiple combinations of travel time and monetary costs. This vignette 
uses a reproducible example to demonstrate how to use `pareto_frontier()` and 
interpret its results.



## 2. What the `pareto_frontier` means.

Imagine a hypothetical journey from A to B. There are multiple route alternatives 
between this origin and destination with varying combinations of travel time and 
cost (figure below).

* Walking from A to B would be the **cheapest** option but it would take 50 minutes.
* The **fastest** option would be to take a bus to a subway station and transfer to the subway. This option would only take 15 minutes, but it would cost $8.
* There are some intermediary alternatives, such as taking:
    * a single bus, $3 for 35 min.
    * two buses with one transfer, $5 for 29 min.
    * taking the subway after walking to the station , $6 for 20 min.

This figure illustrates the Pareto frontier of alternative routes from A to B. In 
other words, it shows the most optimal set of route alternatives between A and B. 
There are certainly other route options, but there is no other option that is both 
faster and cheaper at the same time.

```{r, echo = FALSE, fig.width=7, fig.height=4}
library(ggplot2)

# data.frame
df <- structure(list(option = c(1, 2, 3, 4, 5), 
                     modes = c("Walk", "Bus","Bus + Bus", "Subway", "Bus + Subway"), 
                     time = c(50, 35, 29, 20, 15), 
                     cost = c(0, 3, 5, 6, 8)), class = "data.frame", 
                row.names = c(NA, -5L))

# figure
ggplot(data=df, aes(x=cost, y=time, label = modes)) + 
  geom_step(linetype = "dashed") +
  geom_point() +
  geom_text(color='gray30', hjust = -.2, nudge_x = 0.05, angle = 45) +
  labs(title='Pareto frontier of alternative routes from A to B', subtitle = 'Hypotetical example') +
  scale_x_continuous(name="Travel cost (BRL)", breaks=seq(0,12,3)) +
  scale_y_continuous(name="Travel time (minutes)", breaks=seq(0,60,10)) +
  coord_cartesian(xlim = c(0,14), ylim = c(0, 60)) +
  theme_classic()
```

<br>

This kind of abstraction allows us to have a better grasp of the trade-offs between travel time and monetary cost passengers face when using public transport. It also 
allows us to calculate cumulative-opportunity accessibility metrics with cutoffs 
for both time and cost (e.g. the number of jobs reachable from a given origin with 
limits of 40 minutes and $5) (ref paper by Matt and Anson).

Let's see a couple concrete examples showing how `r5r` can calculate the Pareto 
frontier for multiple origins.



## 3. Demonstration of `pareto_frontier()`.

### 3.1 Build routable transport network with `setup_r5()`

First, let's build the network and create the routing inputs. In this example 
we'll be using the a sample data set for the city of Porto Alegre (Brazil) 
included in `r5r`.

```{r, message = FALSE}
# increase Java memory
options(java.parameters = "-Xmx2G")

# load libraries
library(r5r)
library(data.table)
library(ggplot2)
library(dplyr)

# build a routable transport network with r5r
data_path <- system.file("extdata/poa", package = "r5r")
r5r_core <- setup_r5(data_path)

# routing inputs
mode <- c('walk', 'transit')
max_trip_duration <- 90 # minutes

# load origin/destination points of interest
points <- fread(file.path(data_path, "poa_points_of_interest.csv"))

```


### 3.2 Set up the fare structure

Now we need to set what are the fare rules of our public transport system. These 
rules will be used by `R5` to calculate the monetary cost of alternative routes. 
In the case of Porto Alegre, the fare rules are as follows:
* Each bus ticket costs R$ 4.80.
*  Riding a second bus adds `$` 2.40 to the total cost. Subsequent bus rides cost the full ticket price of $ 4.80.
* Each train ticket costs $ 4.50. Once a passenger enters a train station, she can take an unlimited amount of train trips as long as she doesn’t leave a station.
* The integrated fare between bus and train has a 10% discount, which totals $ 8.37.

We can create `list` object with these fare rules with the support of the `setup_fare_structure()` function as shown in the code below. A detailed explanation of how to use the fare structure of `5r5` can be found in [(this other vignette)](https://ipeagit.github.io/r5r/articles/fare_structure.html).


```{r}
# create basic fare structure
fare_structure <- setup_fare_structure(r5r_core, 
                                       base_fare = 4.8,
                                       by = "MODE")

# update the cost of bus and train fares
fare_structure$fares_per_type[, fare := fcase(type == "BUS", 4.80,
                                             type == "RAIL", 4.50)]

# update the cost of tranfers
fare_structure$fares_per_transfer[, fare := fcase(first_leg == "BUS" & second_leg == "BUS", 7.2,
                                                 first_leg != second_leg, 8.37)]

# update transfer_time_allowance to 60 minutes
fare_structure$transfer_time_allowance <- 60

fare_structure$fares_per_type[type == "RAIL", unlimited_transfers := TRUE]
fare_structure$fares_per_type[type == "RAIL", allow_same_route_transfer := TRUE]

```

For convenience, we can save these fare rules as a `zip` file and load again for 
a future application.


```{r}
# save fare rules to temp file
temp_fares <- tempfile(pattern = "fares_poa", fileext = ".zip")
r5r::write_fare_structure(fare_structure, file_path = temp_fares)


fare_structure <- r5r::read_fare_structure(file.path(data_path, "fares/fares_poa.zip"))
```


### 3.3 Calculating a `pareto_frontier()`.

In this example, we calculate the Pareto frontier from all origins to all 
destinations considering multiple cutoffs of monetary costs:
- $1, which would only allow for walking trips
- $4.5, which would only allow for rail trips
- $4.8, which would allow for a single bus trip
- $7.20, which would allow for bus + bus
- $8.37, which would allow for walking walking + bus + rail

```{r}
departure_datetime <- as.POSIXct("13-05-2019 14:00:00", 
                                 format = "%d-%m-%Y %H:%M:%S")

prtf <- pareto_frontier(r5r_core,
                      origins = points,
                      destinations = points,
                      mode = c("WALK", "TRANSIT"),
                      departure_datetime = departure_datetime,
                      fare_structure = fare_structure,
                      fare_cutoffs = c(1, 4.5, 4.8, 7.20, 8.37),
                      progress = TRUE
                      )
head(prtf)

```


For the sake of illustration, let's check the optimum route alternatives from 
the Farrapos train station to (a) the Praia de Belas shopping mall and (b) the 
Moinhos hospital. An optimum route alternative means that one cannot make a faster trip without 
increasing costs, and one cannot make a cheaper trip without increasing travel time.


```{r, echo = FALSE, fig.width=7, fig.height=4}
# select origin and destinations
pf2 <- dplyr::filter(prtf, to_id == 'farrapos_station'  &
                       from_id %in% c('moinhos_de_vento_hospital', 
                                      'praia_de_belas_shopping_center'))

# recode modes
pf2[, modes := fcase(monetary_cost == 1,   'Walk',
                     monetary_cost == 4.5, 'Train',
                     monetary_cost == 4.8, 'Bus',
                     monetary_cost == 7.2, 'Bus + Bus',
                     monetary_cost == 8.37, 'Bus + Train')]

# plot
ggplot(data=pf2, aes(x=monetary_cost, y=travel_time, color=from_id, label = modes)) + 
  geom_step(linetype = "dashed") +
  geom_point() +
  geom_text(color='gray30', hjust = -.2, nudge_x = 0.05, angle = 45) +
  labs(title='Pareto frontier of alternative routes from Farrapos station to:', 
       subtitle = 'Praia de Belas shopping mall and Moinhos hospital',
       color='Destination') +
  scale_x_continuous(name="Travel cost ($)", breaks=seq(0,12,2)) +
  scale_y_continuous(name="Travel time (minutes)", breaks=seq(0,120,20)) +
  coord_cartesian(xlim = c(0,12), ylim = c(0, 120)) +
  theme_classic() + theme(legend.position=c(.2,0.8))
```

### Cleaning up after usage

`r5r` objects are still allocated to any amount of memory previously set after they are done with their calculations. In order to remove an existing `r5r` object and reallocate the memory it had been using, we use the `stop_r5` function followed by a call to Java's garbage collector, as follows:

```{r, message = FALSE}
r5r::stop_r5(r5r_core)
rJava::.jgc(R.gc = TRUE)
```

If you have any suggestions or want to report an error, please visit [the package GitHub page](https://github.com/ipeaGIT/r5r).


## References
