---
title: 'Using custom OSM car speeds and LTS'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "This vignette shows how to calculate travel times and accessibility with custom OSM car speeds and Level of Traffic Stress, which can be used to simulate different scenarios of traffic congestion and road closures, or interventions in cycling infrastructure."
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Using custom OSM car speeds and LTS} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(tolower(Sys.getenv("NOT_CRAN")), "true"),
  out.width = "100%"
)

```

# 1. Introduction

By default, `R5` routes cars at the legal speed limit of each OSM road edge, a "free flow" scenario without congestion. Real average speeds are usually lower because of traffic and driving behavior. Similarly, `R5` infers the cycling Level of Traffic Stress (LTS) of each road segment from road speeds, road hierarchy and cycling infrastructure.

This vignette shows how to calculate travel times and accessibility with custom OSM car speeds or LTS, to simulate traffic congestion, road closures or interventions in cycling infrastructure. Changes can be passed as (1) a `data.frame`, for individual road segments, or (2) an `sf` object, for roads within / touching its geometries. Examples use the Porto Alegre (Brazil) sample data included in `r5r`.

```{r, message=FALSE}
# increase Java memory
options(java.parameters = "-Xmx2G")

# load libraries
library(r5r)
library(dplyr)
library(data.table)
library(ggplot2)

# data path
data_path <- system.file("extdata/poa", package = "r5r")

# build network
r5r_network <- r5r::build_network(
  data_path = data_path,
  verbose = FALSE
  )
```


# 2. Changing car speeds

All routing and accessibility functions in {r5r} accept custom OSM car speeds through two parameters:

- `new_carspeeds`: here, users must pass either a `data.frame` that indicates the 
new car speed for each OSM edge id, OR an `sf data.frame` polygon that indicates 
the new car speed for all the roads that fall within each polygon.
- `carspeed_scale`: this parameter allows one to set the default car speed for all 
of the road segments not specified in `new_carspeeds`. By default, `carspeed_scale = 1` and the speeds of the unlisted roads are kept unchanged.

## 2.1 Changing car speeds by OSM edge

Here we pass new car speeds in a sample `data.frame` shipped with the package. It must contain the columns `"osm_id"`, `"max_speed"` and `"speed_type"`. `"speed_type"` is character, either `"scale"` or `"km/h"`: whether `"max_speed"` is relative to the original speed (`"scale"`) or an absolute speed (`"km/h"`).

```{r, message=FALSE}
# read data.frame with new car speeds
edge_speed_factors <- read.csv(
  file.path(data_path, "poa_osm_congestion.csv")
  )

head(edge_speed_factors)
```

Here all `"max_speed"` values are `0.5` with `speed_type == "scale"`: the listed OSM edges are driven at 50% of their original OSM speed. Pass `new_carspeeds` and `carspeed_scale` to any routing or accessibility function. Here, `carspeed_scale = 0.8` also slows all unlisted roads to 80% of their speed:

```{r, message=FALSE}
# origins and destination points
points <- read.csv(file.path(data_path, "poa_points_of_interest.csv"))

# travel time matrix
ttm_congestion <- r5r::travel_time_matrix(
  r5r_network = r5r_network,
  origins = points,
  destinations = points,
  mode = 'car',
  departure_datetime = Sys.time(),
  max_trip_duration = 30,
  new_carspeeds = edge_speed_factors,
  carspeed_scale = 0.8
)
```
Obs. Halving speeds does not necessarily double travel times: car travel times also depend on intersections, and slower roads can change the route and hence the trip distance.

### 2.1.1 Setting different congestion levels by road hierarchy

Here we assume congestion is heavier on roads of higher hierarchy. First, read the OSM roads from the `.pbf` file and keep the road types of interest:


```{r, message=FALSE}
# path to OSM pbf
pbf_path <- paste0(data_path, "/poa_osm.pbf")
  
# read layer of lines from pbf
roads <- sf::st_read(
  pbf_path, 
  layer = 'lines', 
  quiet = TRUE
  )

# Filter only road types of interest
rt <- c("motorway", "primary", "secondary", "tertiary") 

roads <- roads |>
  select(osm_id, highway) |>
  filter(highway %in% rt)

head(roads)
```

The selected roads:

```{r}
# map
plot(roads["highway"])

```

Then add a `"max_speed"` column by road type and a `speed_type` column set to `"scale"`, and make `osm_id` numeric:

```{r}
new_edge_speeds <- roads |>
  mutate( 
    osm_id = as.numeric(osm_id),
    max_speed = case_when(
      highway == "motorway"  ~ 0.75,
      highway == "primary"   ~ 0.8,
      highway == "secondary" ~ 0.85,
      highway == "tertiary"  ~ 0.9)) |>
  sf::st_drop_geometry()

new_edge_speeds$speed_type <- "scale"

head(new_edge_speeds)

```

Travel times with the modified car speeds:

```{r, message = FALSE, warning=FALSE}
# travel time matrix
ttm_congestion <- r5r::travel_time_matrix(
  r5r_network = r5r_network,
  origins = points,
  destinations = points,
  mode = 'car',
  departure_datetime = Sys.time(),
  max_trip_duration = 30,
  new_carspeeds = new_edge_speeds
  )

```

### 2.1.2 Setting the same speed limit for the selected road types

To set a 40 km/h speed limit on the same road types (motorway, primary, secondary, tertiary), set `max_speed = 40` and `speed_type = "km/h"` in the previous table:


```{r, message = FALSE}
# edit table with custom speeds to 40 km/h
new_edge_speeds40 <- new_edge_speeds |>
  mutate(max_speed = 40,
         speed_type = "km/h")
  
# travel time matrix
ttm_congestion <- r5r::travel_time_matrix(
  r5r_network = r5r_network,
  origins = points,
  destinations = points,
  mode = 'car',
  departure_datetime = Sys.time(),
  max_trip_duration = 30,
  new_carspeeds = new_edge_speeds40
  )
```

### Extra tip: 

- **Road closure**: one can simulate a road closure by setting the `"max_speed"` value to `0`. This can be quite handy for studies that try to measure the resilience of transport systems to network disruptions.


## 2.2 Changing car speeds with a spatial polygon

Instead of listing OSM edges, you can set the speed of all roads within one or more polygons. The sample data has two polygons: one covering the extended city center, and one covering a few roads that connect two major avenues.

```{r, message = FALSE}
# read sf with congestion polygons
congestion_poly <- readRDS(file.path(data_path, "poa_poly_congestion.rds"))

# preview
mapview::mapview(congestion_poly, zcol="scale")
```

The `sf data.frame` must have these columns:

- `"poly_id"`: a unique id for each polygon
- `"scale"`: the speed scaling factor for each polygon, relative to original speeds. Absolute speeds in km/h are *not* supported.
- `"priority"`: a number ranking which polygon should be considered in case of overlapping polygons.

```{r, message = FALSE}
head(congestion_poly)
```
Here, roads in the city center run at 70% of the speed limit, roads in the second polygon at 80%, and, with `carspeed_scale = 0.95`, all other roads at 95%:

```{r, message = FALSE}
ttm_congestion <- r5r::travel_time_matrix(
  r5r_network = r5r_network,
  origins = points,
  destinations = points,
  mode = 'car',
  departure_datetime = Sys.time(),
  max_trip_duration = 30,
  new_carspeeds = congestion_poly,
  carspeed_scale = 0.95
  )
```

# 3. Changing cycling LTS values

LTS is changed like car speeds: with a `data.frame` of new LTS values per OSM edge id, or with an `sf data.frame`. Here the `sf` must be of type LINESTRING; R5 finds the nearest road to each line and updates its LTS.

## 3.1 Changing LTS by OSM edge

Suppose protected cycle lanes or tracks were built on a few roads. The `data.frame` must contain an `"lts"` column with the new LTS value.

```{r, message=FALSE}
# read data.frame with new lts
edge_lts <- read.csv(
  file.path(data_path, "poa_osm_lts.csv")
  )

head(edge_lts)
```
Pass it to `new_lts` in any routing/accessibility function:

```{r, message = FALSE}
ttm_new_lts <- r5r::travel_time_matrix(
  r5r_network = r5r_network,
  origins = points,
  destinations = points,
  mode = 'bicycle',
  departure_datetime = Sys.time(),
  max_trip_duration = 30,
  new_lts = edge_lts
  )
```

## 3.2. Changing LTS with spatial lines

Alternatively, pass an `sf` LINESTRING of the roads with the intervention. Here, dedicated cycle lanes along all secondary roads:

```{r, message = FALSE}
# read sf with congestion polygons
lts_lines <- readRDS(file.path(data_path, "poa_ls_lts.rds"))

# preview
mapview::mapview(lts_lines, zcol="lts")
```

Pass it to `new_lts` as before:

```{r, message = FALSE}
ttm_new_lts <- r5r::travel_time_matrix(
  r5r_network = r5r_network,
  origins = points,
  destinations = points,
  mode = 'bicycle',
  departure_datetime = Sys.time(),
  max_trip_duration = 30,
  new_lts = lts_lines
  )
```

# Cleaning up after usage

Stop the network and run Java's garbage collector to free the memory it used:

```{r, message = FALSE}
# stop an specific r5r network
r5r::stop_r5(r5r_network)

# or stop all r5r networks at once
r5r::stop_r5()
rJava::.jgc(R.gc = TRUE)
```


If you have any suggestions or want to report an error, please visit [the package GitHub page](https://github.com/ipea/r5r).

