---
title: 'Using the time_window parameter'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "This vignette shows how to use and interpret the `time_window` parameter in r5r."
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Using the time_window parameter} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(tolower(Sys.getenv("NOT_CRAN")), "true"),
  out.width = "100%"
)

```

## 1. Introduction 

### The problem
Travel time and accessibility estimates require a departure time, and they can differ significantly across departure times because public transport service levels vary across the day [@stepniak2019impact]. Even leaving at `10:00am` instead of `10:04am` can change results considerably, depending on when vehicles arrive and how well transfers are coordinated. This relates to the modifiable temporal unit problem (MTUP) [@pereiramtup2019; @levinson2020manual].

The uncertainty is greater when GTFS feeds have a `frequencies.txt` table, because exact vehicle departure times are unknown [@conway2018accounting; @stewart2022half].


### The solution

A common strategy is to compute estimates for multiple departure times over a time window and take the average or median. Repeating the routing for each departure is slow, though.

In `r5r`, functions such as `travel_time_matrix()` and `accessibility()` have a `time_window` parameter that does this in a single call. This vignette shows how to use it and interpret the results.


## 2. How the `time_window` works and how to interpret the results.

With `time_window`, R<sup>5</sup> computes an estimate for a departure every minute from `departure_datetime` until the end of the window. If the GTFS feeds have a `frequencies.txt` table, `draws_per_minute` sets the number of Monte Carlo draws per minute (default 5, i.e. 300 draws in a 60-minute window); without frequencies, departures are deterministic and `draws_per_minute` does not affect results. For the effect of the number of draws on result stability, see Stewart et al [-@stewart2022half].

The result is not a single estimate but a distribution reflecting the uncertainty within the window. The `percentiles` parameter selects which percentiles to return. For example, a 25th-percentile travel time of 15 minutes from A to B means that 25% of trips departing within the window are shorter than 15 minutes.

## 3. Demonstration of `time_window`.

### 3.1 Build routable transport network with `build_network()`

We use the São Paulo (Brazil) sample data included in `r5r`.


```{r, message = FALSE}
# increase Java memory
options(java.parameters = "-Xmx2G")

# load libraries
library(r5r)
library(sf)
library(data.table)
library(ggplot2)
library(dplyr)

# build a routable transport network with r5r
data_path <- system.file("extdata/spo", package = "r5r")
r5r_network <- build_network(data_path)

# routing inputs
mode <- c('walk', 'transit')
max_walk_time <- 30 # minutes
max_trip_duration <- 90 # minutes

# load origin/destination points
points <- fread(file.path(data_path, "spo_hexgrid.csv"))

# departure datetime
departure_datetime = as.POSIXct("13-05-2019 14:00:00", 
                                format = "%d-%m-%Y %H:%M:%S")
```

### 3.2 Accessibility with `time_window`.

Number of schools reachable from each location within 45 minutes (`decay_function = "step"`, `cutoffs = 45`), for departures between 2pm and 3pm (`time_window = 60`):

```{r, message = FALSE}
# estimate accessibility
acc <- r5r::accessibility(
  r5r_network,   
  origins = points,
  destinations = points, 
  opportunities_colnames = 'schools',
  mode = mode,
  max_walk_time = max_walk_time,
  decay_function = "step",
  cutoffs = 45,
  departure_datetime = departure_datetime,
  progress = FALSE,
  time_window = 60,
  percentiles = c(10, 20, 50, 70, 80)
  )

head(acc, n = 10)

```

The output is in long format, one row per origin and percentile, so the first 5 rows refer to the same origin. The 10th percentile counts the schools reachable in at least 10% of departures, the 50th those reachable in at least half of them; so accessibility never increases as the percentile rises. An accessibility of 0 means no school is reachable within 45 minutes.

The plot shows, for each origin (sorted by median), the range of accessibility between the 10th and 80th percentiles, with the median as a dot:

```{r}
# summarize
df <- acc[, .(min_acc = min(accessibility),
              median = accessibility[which(percentile == 50)],
              max_acc = max(accessibility)), by = id]

# plot
ggplot(data=df) +
  geom_linerange(color='gray', alpha=.5, aes(x = reorder(id, median) , 
                      y=median, ymin=min_acc, ymax=max_acc)) +
  geom_point(color='#0570b0', size=.5, aes(x = reorder(id, median), y=median)) +
  labs(y='N. of schools accessible\nby public transport', x='Origins sorted by accessibility',
       title="Accessibility uncertainty between 2pm and 3pm",
       subtitle = 'Upper limit 10% and lower limit 80% of the times') +
  theme_classic() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```



### 3.3 Travel time matrix with `time_window`.

All-to-all travel times for departures between 2pm and 3pm:

```{r, message = FALSE}
# estimate travel time matrix
ttm <- travel_time_matrix(
  r5r_network,   
  origins = points,
  destinations = points,    
  mode = mode,
  max_walk_time = max_walk_time,
  max_trip_duration = max_trip_duration,
  departure_datetime = departure_datetime,
  progress = TRUE,
  time_window = 60,
  percentiles = c(10, 20, 50, 70, 80)
  )

head(ttm, n = 10)

```

Each `travel_time_pXX` column gives the travel time within which XX% of the trips departing between 2pm and 3pm are completed: `travel_time_p10` for the fastest 10%, `travel_time_p50` the median, and so on. An `NA` means fewer than XX% of the trips arrive within `max_trip_duration` (90 minutes).



### 3.4 Expanded travel time matrix with `time_window`.

With `time_window`, `expanded_travel_time_matrix()` returns the fastest route departing at each minute of the window, instead of percentiles. It can be very memory intensive for large data sets and time windows.

```{r}
ettm <- r5r::expanded_travel_time_matrix(
  r5r_network,
  origins = points[1:30,],
  destinations = points[31:61,],    
  mode = mode,
  max_walk_time = max_walk_time,
  max_trip_duration = max_trip_duration,
  departure_datetime = departure_datetime,
  progress = FALSE,
  time_window = 20
  )

head(ettm, n = 10)

```

### 3.5 Detailed itineraries with `time_window`.

`detailed_itineraries()` returns itineraries rather than travel time or accessibility percentiles. With `time_window`, it returns the fastest itinerary: the shortest travel time measured from its own departure, so it may depart and arrive later than others (use `time_window = 1` for the earliest arrival). With `shortest_path = FALSE`, it also returns sub-optimal alternatives.

Its number of draws per minute is fixed at 1: it simulates about one departure per minute of `time_window`, at random seconds (consecutive departures are 30 to 90 seconds apart, and the same departure times are used in every call with the same origin). So a 10-minute `time_window` simulates about 10 departures. Each departure is routed separately, and with the default `suboptimal_minutes = 0` only the itineraries arriving earliest are kept for each departure, even if a slower one has fewer transfers.

*obs.* `detailed_itineraries()` cannot route public transport trips on frequency-based GTFS feeds; use `gtfstools::frequencies_to_stop_times()` to create a suitable feed.



### Cleaning up after usage

Stop the network and run Java's garbage collector to free the memory it used:

```{r, message = FALSE}
r5r::stop_r5(r5r_network)
rJava::.jgc(R.gc = TRUE)
```

If you have any suggestions or want to report an error, please visit [the package GitHub page](https://github.com/ipeaGIT/r5r).

## References
