---
title: 'Using the time_window parameter'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "This vignette shows how to use and interpret the `time_window` parameter in r5r."
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Using the time_window parameter} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(tolower(Sys.getenv("NOT_CRAN")), "true"),
  out.width = "100%"
)

```

## 1. Introduction 

### The problem
To calculate the travel time from A to B, or to calculate the accessibility level at a given location, one has to select a departure time. However, travel time and accessibility estimates can differ significantly at different departure times because of how public transport service levels vary across the day [@stepniak2019impact]. Even a small difference, say leaving at `10:00am` or `10:04am` might importantly change travel time and accessibility estimates depending on when a person departs relative to when a public transport vehicle arrives, and how well transfers are coordinated given a service timetable. This is a very common issue related to the modifiable temporal unit problem (MTUP) [@pereiramtup2019; @levinson2020manual].

This problem gets even more complicated when public transport GTFS feeds have a `frequencies.txt` table. In these cases, we cannot know the exact departure time of vehicles, what creates greater uncertainty for our travel time and accessibility estimates [@conway2018accounting; @stewart2022half].


### The solution

A common strategy to overcome this problem is to calculate travel times and accessibilities at multiple departure times sampled over a time window, and then take the average or median value. Now you may ask, but how many departure times should I use? You might also be thinking  that doing multiple repeated routing analysis can be cumbersome and take a lot of time. Right?

Here is where `r5r` comes in. Both the `travel_time_matrix()`and `accessibility()` functions have a parameter called `time_window`. When this parameter is set, R<sup>5</sup> will automatically compute multiple travel times / accessibility estimates considering multiple departures per minute within the `time_window` selected by the user.

This vignette shows a reproducible example to explain how one can use the `time_window` and interpret the results.


## 2. How the `time_window` works and how to interpret the results.

As mentioned above, when `time_window` is set, R<sup>5</sup> computes multiple travel times / accessibility estimates starting at the specified `departure_datetime` and within the `time_window` selected by the user.  By default, `r5r` will generate one estimate per minute. Nonetheless, users can set a number to the `draws_per_minute` parameter that will change the number of Monte Carlo draws to perform per time window minute. The default value of `draws_per_minute` is 5, which mean 300 draws in a 60 minutes time window, for example. For a detailed discussion on the effect of number of draws on result stability, see Stewart et al [-@stewart2022half].

In this case, there isn't a single estimate of travel time / accessibility, but a distribution of several estimates that reflect the travel time / accessibility uncertainties in the specified time window. To get our heads around so many estimates, we can use the `percentiles` parameter to specify the percentiles of the distribution we are interested in. For example, if we select the 25th travel time percentile and the results show that the travel time estimate between A and B is 15 minutes, this means that 25% of all trips taken between these points within the specified time window are shorter than 15 minutes.


Let's see a couple concrete examples now.

## 3. Demonstration of `time_window`.

### 3.1 Build routable transport network with `setup_r5()`

First, let's build the network and create the routing inputs. In this example we'll be using the a sample data set for the city of São Paulo (Brazil) included in `r5r`.


```{r, message = FALSE}
# increase Java memory
options(java.parameters = "-Xmx2G")

# load libraries
library(r5r)
library(sf)
library(data.table)
library(ggplot2)
library(dplyr)

# build a routable transport network with r5r
data_path <- system.file("extdata/spo", package = "r5r")
r5r_core <- setup_r5(data_path)

# routing inputs
mode <- c('walk', 'transit')
max_walk_time <- 30 # minutes
max_trip_duration <- 90 # minutes

# load origin/destination points
points <- fread(file.path(data_path, "spo_hexgrid.csv"))

# departure datetime
departure_datetime = as.POSIXct("13-05-2019 14:00:00", 
                                format = "%d-%m-%Y %H:%M:%S")
```

ps. Please keep in mind that the `time_window` only affects the results when the GTFS feeds contain a `frequencies.txt` table.



### 3.2 Accessibility with `time_window`.

In this example we calculate the number of schools accessible from each location within a 60-minute time window departing between 2pm and 3pm. In this example we'll be using a cumulative accessibility metric `decay_function = "step"` with a max time threshold of 45 minutes `cutoffs = 45`.

```{r, message = FALSE}
# estimate accessibility
acc <- r5r::accessibility(r5r_core = r5r_core,   
                          origins = points,
                          destinations = points, 
                          opportunities_colnames = 'schools',
                          mode = mode,
                          max_walk_time = max_walk_time,
                          decay_function = "step",
                          cutoffs = 45,
                          departure_datetime = departure_datetime,
                          progress = FALSE,
                          time_window = 30,
                          percentiles = c(10, 20, 50, 70, 80)
                          )

head(acc, n = 10)

```

This output is in long format, so the first 5 rows show the result for the same origin. In this case, we see that in only 10% of the trips departing from that origin between 2pm and 3pm a person would be able to access up to 111 schools. Meanwhile, 50% of the times she would only access 79 schools. By contrast, the accessibility from the other origin shown in the output above is 0, meaning there are no schools accessible from that location given the max travel time of 45 minutes.

We can use a plot like the one below to visualize this uncertainty in how accessibility levels might vary between 2pm and 3pm depending on the departure time within that 60-minute time window.

```{r}
# summarize
df <- acc[, .(min_acc = min(accessibility),
              median = accessibility[which(percentile == 50)],
              max_acc = max(accessibility)), by = id]

# plot
ggplot(data=df) +
  geom_linerange(color='gray', alpha=.5, aes(x = reorder(id, median) , 
                      y=median, ymin=min_acc, ymax=max_acc)) +
  geom_point(color='#0570b0', size=.5, aes(x = reorder(id, median), y=median)) +
  labs(y='N. of schools accessible\nby public transport', x='Origins sorted by accessibility',
       title="Accessibility uncertainty between 2pm and 3pm",
       subtitle = 'Upper limit 10% and lower limit 80% of the times') +
  theme_classic() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```



### 3.3 Travel time matrix with `time_window`.

Now let's calculate all-to-all travel time estimates within a 60-minute time window departing between 2pm and 3pm and see how the output looks like.

```{r, message = FALSE}
# estimate travel time matrix
ttm <- travel_time_matrix(r5r_core = r5r_core,   
                          origins = points,
                          destinations = points,    
                          mode = mode,
                          max_walk_time = max_walk_time,
                          max_trip_duration = max_trip_duration,
                          departure_datetime = departure_datetime,
                          progress = TRUE,
                          time_window = 20,
                          percentiles = c(10, 20, 50, 70, 80)
                          )

head(ttm, n = 10)

```

Now let's look at the 2nd row of the output above. This output tell us that only 10% of the trips between 2pm and 3pm for that origin-destination pair took 39 minutes or less. Meanwhile, 50% of those trips took up tp 45 minutes and 80% of them were 48-minute long or shorter.

The last row in the result above has a few `NA`s. This tell us that at least 50% of all simulated trips between 2pm and 3pm for that origin-destination pair could not be completed because they took longer than the `max_trip_duration` we have set (90 minutes).



### 3.4 Expanded travel time matrix with `time_window`.

Finally, we can also use the `time_window` in the `expanded_travel_time_matrix()` function. In this case, though, when the user sets a `time_window` value, the `expanded_travel_time_matrix()` will return the fastest route alternative departing each minute within the specified time window. Please note this function can be very memory intensive for large data sets and time windows.

```{r}
ettm <- r5r::expanded_travel_time_matrix(r5r_core = r5r_core,   
                                    origins = points[1:30,],
                                    destinations = points[31:61,],    
                                    mode = mode,
                                    max_walk_time = max_walk_time,
                                    max_trip_duration = max_trip_duration,
                                    departure_datetime = departure_datetime,
                                    progress = FALSE,
                                    time_window = 20)

head(ettm, n = 10)

```


### Cleaning up after usage

`r5r` objects are still allocated to any amount of memory previously set after they are done with their calculations. In order to remove an existing `r5r` object and reallocate the memory it had been using, we use the `stop_r5` function followed by a call to Java's garbage collector, as follows:

```{r, message = FALSE}
r5r::stop_r5(r5r_core)
rJava::.jgc(R.gc = TRUE)
```

If you have any suggestions or want to report an error, please visit [the package GitHub page](https://github.com/ipeaGIT/r5r).

## References
