---
title: 'Travel time matrices'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
abstract: "This vignette shows how to use the travel_time_matrix() and expanded_travel_time_matrix() functions in r5r."
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Travel time matrices} 
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = identical(tolower(Sys.getenv("NOT_CRAN")), "true"),
  out.width = "100%"
)

## removes files previously created by 'build_network()'
#data_path <- system.file("extdata/poa", package = "r5r")
#existing_files <- list.files(data_path)
#files_to_keep <- c(
#  "poa_hexgrid.csv",
#  "poa_osm.pbf",
#  "poa_points_of_interest.csv",
#  "poa_eptc.zip",
#  "poa_trensurb.zip",
#  'fares'
#  )
#files_to_remove <- existing_files[! existing_files %in% files_to_keep]
#invisible(file.remove(file.path(data_path, files_to_remove)))
```

# 1. Introduction 

Many transport planning and modeling tasks need travel time estimates between origins and destinations. `R5` computes realistic door-to-door travel times in multimodal networks very fast, and `r5r` offers three functions for it:

- `travel_time_matrix()`
- `expanded_travel_time_matrix()`
- `arrival_travel_time_matrix()`

This vignette shows, with a reproducible example, how they work and how they differ.


# 2. Build routable transport network with `build_network()`

We use the Porto Alegre (Brazil) sample data included in `r5r`.


```{r, message = FALSE}
# increase Java memory
options(java.parameters = "-Xmx2G")

# load libraries
library(r5r)
library(data.table)
library(ggplot2)

# build a routable transport network with r5r
data_path <- system.file("extdata/poa", package = "r5r")
r5r_network <- build_network(data_path)

# routing inputs
mode <- c('walk', 'transit')
max_trip_duration <- 60 # minutes

# departure time
departure_datetime <- as.POSIXct("13-05-2019 14:00:00", 
                                 format = "%d-%m-%Y %H:%M:%S")

# load origin/destination points
points <- fread(file.path(data_path, "poa_points_of_interest.csv"))

```


# 3. The `travel_time_matrix()` function

`travel_time_matrix()` quickly computes travel times between all origin/destination pairs for a departure time and transport mode. Other parameters include:

- `max_trip_duration`: maximum trip duration
- `max_rides`: maximum number of public transport rides
- `max_walk_time` and `max_bike_time`: maximum walking or cycling time to and from public transport
- `walk_speed` and `bike_speed`: average walking or cycling speed (km/h)
- `max_fare`: maximum monetary cost in public transport. [See this vignette](https://ipea.github.io/r5r/articles/fare_structure.html).


```{r, message = FALSE}
# estimate travel time matrix
ttm <- travel_time_matrix(
  r5r_network,   
  origins = points,
  destinations = points,    
  mode = mode,
  max_trip_duration = max_trip_duration,
  departure_datetime = departure_datetime
  )

head(ttm, n = 10)

```

Travel times can vary significantly across the day with public transport service levels. The `time_window` and `percentiles` parameters handle this efficiently: R<sup>5</sup> computes travel times for departures every minute within `time_window` and returns the selected percentiles ([time window vignette](https://ipea.github.io/r5r/articles/time_window.html)).


# 4. The `expanded_travel_time_matrix()` function

`expanded_travel_time_matrix()` returns more than the total travel time: by default, it also lists the public transport routes taken between each origin/destination pair. With `breakdown = TRUE`, it adds each trip's number of public transport rides and its access, waiting, in-vehicle, transfer and egress times, which can be slower for large data sets.

*A general call to expanded_travel_time_matrix()*
```{r, message = FALSE}
ettm <- expanded_travel_time_matrix(
  r5r_network,   
  origins = points,
  destinations = points,    
  mode = mode,
  max_trip_duration = max_trip_duration,
  departure_datetime = departure_datetime
  )

head(ettm, n = 10)
```

*Calling expanded_travel_time_matrix() with `breakdown = TRUE`*

```{r, message = FALSE}
ettm2 <- expanded_travel_time_matrix(
  r5r_network,   
  origins = points,
  destinations = points,    
  mode = mode,
  max_trip_duration = max_trip_duration,
  departure_datetime = departure_datetime,
  breakdown = TRUE
  )

head(ettm2, n = 10)
```

Over its `time_window` (10 minutes by default), `expanded_travel_time_matrix()` returns the fastest route departing at each minute. This can be very memory intensive for large data sets and time windows.

```{r, message = FALSE}
ettm_window <- expanded_travel_time_matrix(
  r5r_network,   
  origins = points,
  destinations = points,    
  mode = mode,
  max_trip_duration = max_trip_duration,
  departure_datetime = departure_datetime,
  breakdown = TRUE,
  time_window = 10
  )

ettm_window[15:25,]
```



# 5. The `arrival_travel_time_matrix()` function

`travel_time_matrix()` and `expanded_travel_time_matrix()` take a **departure** time. To route by **arrival** time, use `arrival_travel_time_matrix()`: given the latest acceptable arrival time and a maximum trip duration, it returns the travel time of the trip with the latest departure that still arrives in time.

This models trips where arriving by a set time matters, such as getting to work or school by 9 a.m.: people usually take the latest departure that gets them there on time, not the fastest trip that leaves them waiting at the destination.

As in `expanded_travel_time_matrix()`, the output has additional columns (and `breakdown = TRUE` is available).

*A general call to arrival_travel_time_matrix()*
```{r, message = FALSE}

arrival_datetime <- as.POSIXct(
 "13-05-2019 14:00:00",
 format = "%d-%m-%Y %H:%M:%S"
)

arrival_ttm <- arrival_travel_time_matrix(
  r5r_network,
  origins = points,
  destinations = points,
  mode = c("WALK", "TRANSIT"),
  arrival_datetime = arrival_datetime,
  max_trip_duration = 60
)

head(arrival_ttm, n = 10)
```



### Cleaning up after usage

Stop the network and run Java's garbage collector to free the memory it used:

```{r, message = FALSE}
r5r::stop_r5(r5r_network)
rJava::.jgc(R.gc = TRUE)
```

```{r, eval = TRUE, include = FALSE, message = FALSE}
# clean cache (CRAN policy)
r5r::r5r_cache(delete_file = 'all')

```

If you have any suggestions or want to report an error, please visit [the package GitHub page](https://github.com/ipea/r5r).

## References
