---
title: 'r5r: Rapid Realistic Routing with R5 in R'
author: "   "
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: default
  github_document: default
abstract: " write abstract"
urlcolor: blue
vignette: |
  %\VignetteIndexEntry{r5r-intro} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

#> Examples of similar papers
#> - https://transportfindings.org/article/8416-accessibility-toolbox-for-r-and-arcgis
#> 
#> - https://transportfindings.org/article/6945-dodgr-an-r-package-for-network-flow-aggregation
#> 
#> - https://joss.theoj.org/papers/10.21105/joss.01926


# RESEARCH QUESTIONS AND HYPOTHESES 

Transport routing is the process of finding the alternative routes and costs 
that connect places on a given transport network. Routing often involves complex
mathematical algorithms and it is a fundamental step required for transport
accessibility analysis, fleet allocation and transport simulation and planning
more broadly. Despite the long-standing importance of routing analysis to the 
field, transport researchers and practitioners still frequently face practical 
challenges with this task. These challenges involve for example the monetary 
costs of licensed software, data availability issues for data hungry models, and
the long computation times to run multiple routing scenarios, particularly in 
large and complex multimodal transport networks. While there is a growing number
of open source routing models (ref: stplnr, otp, dodgr), most options available 
only return the fastest route option between a set of origins and destinations. 
This simple approach overlooks how often people's actual route choices deviate 
from shortest travel time paths due to imperfect information, monetary costs and
safety issues among other reasons (refs, 1 2 3).
Refs:
1 - https://tram.mcgill.ca/Research/Publications/Who_will_use_it.pdf
2 - https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0134322
3 - https://www.tandfonline.com/doi/abs/10.1080/13658816.2019.1605075

Test [@http://zotero.org/users/local/FtLgDWw9/items/8CHCA2S3]


This paper presents [r5r](https://ipeagit.github.io/r5r/), a new open source 
R package for rapid realistic routing on multimodal transport networks based on
the [Rapid Realistic Routing on Real-world and Reimagined networks (R<sup>5</sup>)](https://github.com/conveyal/r5). R<sup>5</sup> is a powerful 
next-generation routing engine written in Java and developed at Conveyal (ref 1-2)
to replace its predecessor OpenTripPlanner. The r5r package provides a simple 
and friendly interface to run R<sup>5</sup> locally from within R, what allows 
users to efficiently calculate travel time matrices or generate multiple route 
alternatives between origins and destinations using seamless parallel computing.
1 - https://repository.asu.edu/items/54162
2- https://www.jtlu.org/index.php/jtlu/article/view/1074




# METHODS AND DATA

r5r has low data requirements, enabling routing analysis on 
different contexts wherever there is open data on street networks from 
OpenStreetMap (OSM). The package is also easily scalable allowing for fast 
computations at either the city or country level analysis. For multimodal 
transport routing, r5r will automatically combine OpenStreetMap with one or 
multiple public transport data sets in the stantard GTFS format to create a routable
transport network.

The r5r package has 3 fundamental functions. The `setup_r5()` function builds a 
multimodal transport network used for routing in R5. This function (1) 
downloads/updates a compiled JAR file of R5 and stores it locally in the r5r
package directory for future use; and (2) combines the OSM and GTFS data sets to
build a routable network object. For fast routing analysis, r5r currently has 
two functions:

- `travel_time_matrix`: a simple and fast function to efficiently compute travel
time estimates between one or multiple origin/destination pairs. The origin/destination
input can be either a spatial `sf POINT` object, or a `data.frame` containing the
columns `id`, `lon`, `lat.` 

- `detailed_itineraries`:  function that provides the shortest or 
multiple alternative routes between origin/destination pairs. The output also 
brings detailed information for each route alternative at the trip segment level,
including the transport mode, waiting times, travel time and distance of each 
trip segment.

Both routing functions are versatile so users can easily can set customized inputs
such as transport mode, departure date and times, walking  an cycling speeds, 
maximum trip duration, walking distances or number of public transport transfers
etc.





# FINDINGS

Section to demonstrate how to use the core functions of the package. We can use 
our intro vignette as a starting point and build from there with more the analysis
of larger data sets, include couple more data visualization of results and emphasize
the results of multiple route alternatives.


To illustrate functionality, the package includes a small sample data for the 
city of Porto Alegre (Brazil). It includes three files:
- An Open Street Map network in `.pbf` format (*mandatory*)
- A public transport network in `GTFS.zip` format
- The spatial coordinates of the origin destination pairs in a `.csv` file.



```{r, message = FALSE}
# Load packages
library(r5r)
library(sf)
library(data.table)
library(dplyr)
library(ggplot2)

```


```{r}
path <- system.file("extdata", package = "r5r")
list.files(path)

points <- fread(system.file("extdata/poa_hexgrid.csv", package = "r5r"))
points <- points[ c(sample(1:nrow(points), 10, replace=TRUE)), ]
head(points)

```

By default, R allocates only 512MB of memory for Java processes, which is not enough for large queries using r5r. To increase available memory to 4GB, for example, we need to set the java.parameters option at the beginning of the script, as follows:

```{r, message = FALSE}
options(java.parameters = "-Xmx4G")
```



### Build routable transport network

The first step is to build the multimodal transport used for routing in R5. This
is done with the `setup_r5` function. It does two things: (1) downloads a compiled JAR file
of R5 and store it locally in the r5r package directory; (2) combines the osm.pbf and gtfs.zip data sets to build a routable network object. The function 

```{r, message = FALSE}
# Indicate the path where OSM and GTFS data are stored
r5r_core <- setup_r5(data_path = path)

```

**r5** has two core routing functions.

- `detailed_itineraries`
   * Returns a `data.frame sf LINESTRINGs` with one or multiple alternative routes
   between one or multiple origin destination pairs. The data output brings 
   detailed information on transport mode, travel time, walk distance etc for 
   each trip section
 
- `travel_time_matrix`
   * Fast function that returns a simple 'data.frame' with travel time 
   estimates between one or multiple origin destination pairs.
   
   
## Fast many to many travel time matrix

```{r, message = FALSE}
mode <- c("WALK", "TRANSIT")
max_walk_dist <- Inf
max_trip_duration <- 120L
departure_datetime <- as.POSIXct("13-03-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")

# estimate travel time matrix
ttm <- travel_time_matrix(r5r_core,
                          origins = points,
                          destinations = points,
                          mode,
                          departure_datetime,
                          max_walk_dist,
                          max_trip_duration)

head(ttm)
```

## Detailed itineraries

Allows one to get the fastest or multiple alternative routes between a set of
origins and destinations.

Create simple example here and plot the result below

### visualize results


```{r, message = FALSE}
# extract OSM network
street_net <- street_network_to_sf(r5r_core)

# plot
ggplot() +
        geom_sf(data = street_net$edges, color='gray85') +
        theme_minimal()

```












# Acknowledgments

Conveyal for R5


# References




#> Rafa's notes
#> jean: #>-  mass of assumptions regarding decision-making and routing (Miller, 2018). 
#>Haugen, 2011; Schwanen, 2008
#>
#>Miller, E.J., 2018. Accessibility: measurement and application in transportation planning. Transp. Rev. 38, #>551–555. https://doi.org/10.1080/01441647.2018.1492778
#>
#>This article presents **r5r**, a new open-source `R` package for rapid 
#>realistic routing on multimodal transport networks.
