---
title: 'r5r: Rapid Realistic Routing with R5 in R'
author: "Rafael H. M. Pereira, Marcus Saraiva, Daniel Herszenhut, Carlos Kaue Braga"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: default
  github_document: default
abstract: "write abstract"
urlcolor: blue
vignette: |
  %\VignetteIndexEntry{r5r-intro} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

set.seed(0)
```

<!-- Examples of similar papers -->
<!--  - https://transportfindings.org/article/8416-accessibility-toolbox-for-r-and-arcgis -->

<!--  - https://transportfindings.org/article/6945-dodgr-an-r-package-for-network-flow-aggregation -->

<!--  - https://joss.theoj.org/papers/10.21105/joss.01926 -->

dasdasdds

# RESEARCH QUESTIONS AND HYPOTHESES 

Transport routing is the process of finding the alternative routes and costs 
that connect places on a given transport network. Routing often involves complex
mathematical algorithms and it is a fundamental step required for transport
accessibility analysis, fleet allocation and transport simulation and planning
more broadly. Despite the long-standing importance of routing analysis to the 
field, transport researchers and practitioners still frequently face practical 
challenges with this task. These challenges involve for example the monetary 
costs of licensed software, data availability issues for data hungry models, and
the long computation times to run multiple routing scenarios, particularly in 
large and complex multimodal transport networks. While there is a growing number
of open source routing models [@opentripplanneropentripplanner; 
@lovelace2019stplanr; @padgham2019dodgr], most options available only return the
fastest route option between a set of origins and destinations. This simple
approach overlooks how often people's actual route choices deviate from shortest
travel time paths due to imperfect information, monetary costs and safety issues
among other reasons [@larsen2011travel; @zhu2015people; @conway2019getting].

This paper presents [r5r](https://ipeagit.github.io/r5r/), a new open source 
R package for rapid realistic routing on multimodal transport networks based on
the [Rapid Realistic Routing on Real-world and Reimagined networks (R<sup>5</sup>)](https://github.com/conveyal/r5). 
R<sup>5</sup> is a powerful next-generation routing engine written in Java and
developed at Conveyal [@conway2017evidencebased; @conway2018accounting] to
replace its predecessor OpenTripPlanner. The r5r package provides a simple and
friendly interface to run R<sup>5</sup> locally from within R, what allows users
to efficiently calculate travel time matrices or generate multiple route 
alternatives between origins and destinations using seamless parallel computing.

# METHODS AND DATA

r5r has low data requirements, enabling routing analysis on different contexts 
wherever there is open data on street networks from OpenStreetMap (OSM). The 
package is also easily scalable allowing for fast computations at either the 
city or country level analysis. For multimodal transport routing, r5r will 
automatically combine OSM data with one or multiple public transport data sets 
in the standard GTFS format to create a routable transport network.

The r5r package has 3 fundamental functions. The `setup_r5()` function builds a 
multimodal transport network used for routing in R<sup>5</sup>. This function 
(1) downloads/updates a compiled R<sup>5</sup> JAR file, tailored for the 
purposes of the package, and stores it locally in the r5r package directory for 
future use; and (2) combines the OSM and GTFS data sets to build a routable 
network object. For fast routing analysis, r5r currently has two functions:

- `travel_time_matrix()`: used to compute travel time estimates between one or 
multiple origin/destination pairs. The origin/destination input can be either a 
spatial `sf POINT` object, or a `data.frame` containing the columns `id`, `lon`, 
`lat.` 

- `detailed_itineraries()`:  used to compute detailed information at the trip 
segment level on either the shortest or multiple alternative routes between one
or multiple origin/destination pairs. Includes information such as the transport 
mode, waiting time, travel time and distance.

Both routing functions are versatile so users can easily set customized inputs
such as transport mode, departure date and times, walking and cycling speeds, 
maximum trip duration, walking distances and number of public transport 
transfers, etc.

# FINDINGS

The package can be installed with the following command:

```{r, message = FALSE, eval = FALSE}
devtools::install_github("ipeaGIT/r5r", subdir = "r-package")
```

The first step when using r5r is to set the amount of memory available to Java. 
By default `R` allocates only 512 MB of memory for Java processes, which is not 
enough for large queries using r5r. To increase available memory to 2GB, for 
example, we need to set the `java.parameters` option at the beginning of the 
script as follows:

```{r, message = FALSE}
options(java.parameters = "-Xmx2G")
```

Please note that `R` only allocates memory to Java once during a running 
session. If a Java process has been initiated before, either by r5r or any other 
Java-dependent package or function, it might be required to restart the current 
`R` session in order to effectively change the amount of memory allocated.

The package can be attached (with some other packages used in this article) with 
the following command:

```{r, message = FALSE}
library(r5r)
library(sf)
library(data.table)
library(ggplot2)
```

To illustrate functionality, the package includes a small sample data set for 
the city of Porto Alegre (Brazil). It includes three files:

- An OSM network in `.pbf` format (*mandatory*).
- A public transport network in `GTFS.zip` format.
- The spatial coordinates of points covering the area in `.csv` format.

The `.pbf` and `GTFS.zip` files are used to build the transport network. While 
the first is mandatory, since it is required to create a street network, the 
latter is only used to calculate public transport trips, thus may not be 
necessary if that's not intended.

```{r}
data_path <- system.file("extdata", package = "r5r")
list.files(data_path)

points <- fread(system.file("extdata/poa_hexgrid.csv", package = "r5r"))
points <- points[c(sample(1:nrow(points), 100))]
head(points)
```


### Build routable transport network

The first step is to build the multimodal transport used for routing in R5. This
is done with the `setup_r5` function. It does two things: (1) downloads a compiled JAR file
of R5 and store it locally in the r5r package directory; (2) combines the osm.pbf and gtfs.zip data sets to build a routable network object. The function 

```{r, message = FALSE}
# Indicate the path where OSM and GTFS data are stored
r5r_core <- setup_r5(data_path)
```

**r5** has two core routing functions.

- `detailed_itineraries`
   * Returns a `data.frame sf LINESTRINGs` with one or multiple alternative routes
   between one or multiple origin destination pairs. The data output brings 
   detailed information on transport mode, travel time, walk distance etc for 
   each trip section
 
- `travel_time_matrix`
   * Fast function that returns a simple 'data.frame' with travel time 
   estimates between one or multiple origin destination pairs.
   
   
## Fast many to many travel time matrix

```{r, message = FALSE}
mode <- c("WALK", "TRANSIT")
max_walk_dist <- Inf
max_trip_duration <- 120L
departure_datetime <- as.POSIXct("13-03-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")

# estimate travel time matrix
ttm <- travel_time_matrix(r5r_core,
                          origins = points,
                          destinations = points,
                          mode,
                          departure_datetime,
                          max_walk_dist,
                          max_trip_duration)

head(ttm)
```

## Detailed itineraries

Allows one to get the fastest or multiple alternative routes between a set of
origins and destinations.

Create simple example here and plot the result below

### visualize results


```{r, message = FALSE}
# extract OSM network
street_net <- street_network_to_sf(r5r_core)

# plot
ggplot() +
   geom_sf(data = street_net$edges, color='gray85') +
   theme_minimal()

```












# Acknowledgments

Conveyal for R5


# References




<!-- Rafa's notes -->
<!-- jean: #>-  mass of assumptions regarding decision-making and routing (Miller, 2018).  -->
<!-- Haugen, 2011; Schwanen, 2008 -->

<!-- Miller, E.J., 2018. Accessibility: measurement and application in transportation planning. Transp. Rev. 38, #>551–555. https://doi.org/10.1080/01441647.2018.1492778 -->

<!-- This article presents **r5r**, a new open-source `R` package for rapid  -->
<!-- realistic routing on multimodal transport networks. -->
