---
title: 'r5r: Rapid Realistic Routing with R5 in R'
author: "Rafael H. M. Pereira, Marcus Saraiva, Daniel Herszenhut, Carlos Kaue Braga"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: default
  github_document: default
abstract: "write abstract"
urlcolor: blue
vignette: |
  %\VignetteIndexEntry{r5r-intro} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

set.seed(0)
```

<!-- Examples of similar papers -->
<!--  - https://transportfindings.org/article/8416-accessibility-toolbox-for-r-and-arcgis -->

<!--  - https://transportfindings.org/article/6945-dodgr-an-r-package-for-network-flow-aggregation -->

<!--  - https://joss.theoj.org/papers/10.21105/joss.01926 -->

# RESEARCH QUESTIONS AND HYPOTHESES 

Transport routing is the process of finding the alternative routes and costs 
that connect places on a given transport network. Routing often involves complex
mathematical algorithms and it is a fundamental step required for transport
accessibility analysis, fleet allocation and transport simulation and planning
more broadly. Despite the long-standing importance of routing analysis to the 
field, transport researchers and practitioners still frequently face practical 
challenges with this task. These challenges involve for example the monetary 
costs of licensed software, data availability issues for data hungry models, and
the long computation times to run multiple routing scenarios, particularly in 
large and complex multimodal transport networks. While there is a growing number
of open source routing models [@opentripplanneropentripplanner; 
@lovelace2019stplanr; @padgham2019dodgr], most options available only return the
fastest route option between a set of origins and destinations. This simple
approach overlooks how often people's actual route choices deviate from shortest
travel time paths due to imperfect information, monetary costs and safety issues
among other reasons [@larsen2011travel; @zhu2015people; @conway2019getting].

This paper presents [r5r](https://ipeagit.github.io/r5r/), a new open source 
R package for rapid realistic routing on multimodal transport networks based on
the [Rapid Realistic Routing on Real-world and Reimagined networks (R<sup>5</sup>)](https://github.com/conveyal/r5). 
R<sup>5</sup> is a powerful next-generation routing engine written in Java and
developed at Conveyal [@conway2017evidencebased; @conway2018accounting] to
replace its predecessor OpenTripPlanner. The r5r package provides a simple and
friendly interface to run R<sup>5</sup> locally from within R, what allows users
to efficiently calculate travel time matrices or generate multiple route 
alternatives between origins and destinations using seamless parallel computing.

# METHODS AND DATA

r5r has low data requirements, enabling routing analysis on different contexts 
wherever there is open data on street networks from OpenStreetMap (OSM). The 
package is also easily scalable allowing for fast computations at either the 
city or country level analysis. For multimodal transport routing, r5r will 
automatically combine OSM data with one or multiple public transport data sets 
in the standard GTFS format to create a routable transport network.

The r5r package has 3 fundamental functions. The `setup_r5()` function builds a 
multimodal transport network used for routing in R<sup>5</sup>. This function 
(1) downloads/updates a compiled R<sup>5</sup> JAR file, tailored for the 
purposes of the package, and stores it locally in the r5r package directory for 
future use; and (2) combines the OSM and GTFS data sets to build a routable 
network object. For fast routing analysis, r5r currently has two functions:

- `travel_time_matrix()`: used to compute travel time estimates between one or 
multiple origin/destination pairs. The origin/destination input can be either a 
spatial `sf POINT` object, or a `data.frame` containing the columns `id`, `lon`, 
`lat.` 

- `detailed_itineraries()`:  used to compute detailed information at the trip 
segment level on either the shortest or multiple alternative routes between one
or multiple origin/destination pairs. Includes information such as the transport 
mode, waiting time, travel time and distance.

Both routing functions are versatile so users can easily set customized inputs
such as transport mode, departure date and times, walking and cycling speeds, 
maximum trip duration, walking distances and number of public transport 
transfers, etc.

# FINDINGS

The package can be installed with the following commands:

```{r, message = FALSE, eval = FALSE}
# stable version
install.packages("r5r")

# development version
devtools::install_github("ipeaGIT/r5r", subdir = "r-package")
```

The first step when using r5r is to set the amount of memory available to Java. 
By default `R` allocates only 512 MB of memory for Java processes, which is not 
enough for large queries using r5r. To increase available memory to 2GB, for 
example, we need to set the `java.parameters` option at the beginning of the 
script as follows:

```{r, message = FALSE}
options(java.parameters = "-Xmx2G")
```

Please note that `R` only allocates memory to Java once during a running 
session. If a Java process has been initiated before, either by r5r or any other 
Java-dependent package or function, it might be required to restart the current 
`R` session in order to effectively change the amount of memory allocated.

The package can be attached (with some other packages used in this article) with 
the following command:

```{r, message = FALSE}
library(r5r)
library(sf)
library(data.table)
library(ggplot2)
```

To illustrate functionality, the package includes a small sample data set for 
the city of Porto Alegre (Brazil). It includes three files:

- An OSM network in `.pbf` format.
- A public transport network in `GTFS.zip` format.
- The spatial coordinates of points covering the area in `.csv` format.

```{r}
data_path <- system.file("extdata", package = "r5r")
list.files(data_path)

points <- fread(system.file("extdata/poa_hexgrid.csv", package = "r5r"))
points <- points[c(sample(1:nrow(points), 100))]
head(points)
```

### Build a routable transport network

Building a transport network with r5r is as simple as a one command line:

```{r, message = FALSE}
r5r_core <- setup_r5(data_path)
```

The function uses the `.pbf` and the `GTFS.zip` files in the directory pointed 
by `data_path` to create a multimodal transport network used for routing by 
R<sup>5</sup>. The resulting `network.dat`, as well as some other files used by 
R<sup>5</sup>, is saved inside the supplied directory.

While the presence of a `.pbf` file is mandatory, since it is required to create 
a street network, the `GTFS.zip` file is only used to calculate public transport 
trips, thus may not be necessary if that's not intended. Multiple GTFS can be 
used simultaneously, in which case all files are merged into a single one 
automatically.
   
### Calculate a travel time matrix

Calculating a travel time matrix with r5r can be done very efficiently with the
`travel_time_matrix()` function. The function takes travel restrictions and
origins/destinations spatial location as inputs and outputs a `data.table` with 
the travel time between all combinations of origins and destinations that can be 
reached with the given restrictions:

```{r, message = FALSE}
# set restrictions
mode <- c("WALK", "TRANSIT")
max_walk_dist <- 1000
max_trip_duration <- 120
departure_datetime <- as.POSIXct("13-03-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")

# calculate travel time matrix
ttm <- travel_time_matrix(r5r_core,
                          origins = points,
                          destinations = points,
                          mode,
                          departure_datetime,
                          max_walk_dist,
                          max_trip_duration)

ttm
```

Note that the resulting `ttm` object only holds the travel time of 1043 
origin/destination pairs, even though the `points` data set passed both to 
`origins` and `destinations` argument contains 100 points. This is because out 
of the total 10000 possible trips between the supplied points, only 1043 could 
be completed within the specified restrictions. Removing any walking distance 
limits (by setting it to infinite) and increasing maximum trip duration 
accordingly results in a "complete" output:

```{r, message = FALSE}
# set softer restrictions
max_walk_dist <- Inf
max_trip_duration <- 360

# calculate travel time matrix
ttm <- travel_time_matrix(r5r_core,
                          origins = points,
                          destinations = points,
                          mode,
                          departure_datetime,
                          max_walk_dist,
                          max_trip_duration)

ttm
```

### Calculate detailed itineraries

Often enough one wants more information on the itineraries between 
origin/destination pairs than simply the travel time. For such cases, the 
package includes the `detailed_itineraries()` function, which also takes travel 
restrictions as inputs, but outputs a much more detailed data set at the trip 
segment level. Another key feature of this function is that it (optionally) 
provides multiple alternative routes between origin/destination pairs:

```{r, message = FALSE}
# load points of interests and set origin and destination
points <- read.csv(file.path(data_path, "poa_points_of_interest.csv"))
origins <- points[c(10, 14),]
destinations <- points[c(12, 15),]

# set restrictions
mode <- c("WALK", "TRANSIT")
max_walk_dist <- Inf
max_trip_duration <- 120
departure_datetime <- as.POSIXct("13-03-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")

dit <- detailed_itineraries(r5r_core,
                            origins,
                            destinations,
                            mode,
                            departure_datetime,
                            max_walk_dist,
                            shortest_path = FALSE)

dit
```

Since the `shortest_path` argument has been set to `FALSE`, `dit` holds 
information for multiple itineraries between the same origin/destination pairs:

```{r, message = FALSE}
setDT(copy(dit))[, .(n_itineraries = uniqueN(option)), by = .(fromId, toId)]
```

Had `shortest_path` been set to `TRUE`, only the shortest path would have been 
returned:

```{r, message = FALSE}
dit_short <- detailed_itineraries(r5r_core,
                                  origins,
                                  destinations,
                                  mode,
                                  departure_datetime,
                                  max_walk_dist,
                                  shortest_path = TRUE)

setDT(copy(dit_short))[, .(n_itineraries = uniqueN(option)), by = .(fromId, toId)]
```

### Visualize results

`detailed_itineraries()` output is a `sf` object, which means that each trip 
segment is associated to a spatial geometry that can be plotted with common `R` 
packages, such as `ggplot2`. The package also offers a handy 
`street_network_to_sf()` function which extracts the OSM street network used by 
R<sup>5</sup> as a `sf` object and that can be used to provide geographic 
context for the visualization.

```{r, message = FALSE, fig.width = 7, fig.height = 5}
# extract OSM network
street_net <- street_network_to_sf(r5r_core)

dit$facet_labels <- paste0(dit$fromId, " to\n", dit$toId)

ggplot() +
   geom_sf(data = street_net$edges, color = "gray85") +
   geom_sf(data = dit, aes(color = mode), size = 1) +
   facet_wrap(~ facet_labels, nrow = 1) +
   theme_minimal() + 
   theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))
```

The picture above can be used to compare the walking-only to the walking and 
transit itineraries between each origin/destination pair. Given that the sample 
`GTFS.zip` file includes only very few public transport routes, the resulting 
plot picture some odd scenarios, such as a fairly lengthy circular-like bus 
route, in the leftmost facet, and a very long walking access trip to the bus, 
in the rightmost facet.  

# Acknowledgments

The [R<sup>5</sup> routing engine](https://github.com/conveyal/r5) is developed 
at [Conveyal](https://www.conveyal.com/) with contributions from several people. 
This work was supported by the Brazilian Institute for Applied Economic Research 
(Ipea).

# References




<!-- Rafa's notes -->
<!-- jean: #>-  mass of assumptions regarding decision-making and routing (Miller, 2018).  -->
<!-- Haugen, 2011; Schwanen, 2008 -->

<!-- Miller, E.J., 2018. Accessibility: measurement and application in transportation planning. Transp. Rev. 38, #>551–555. https://doi.org/10.1080/01441647.2018.1492778 -->

<!-- This article presents **r5r**, a new open-source `R` package for rapid  -->
<!-- realistic routing on multimodal transport networks. -->
