---
title: 'r5r: rapid realistic routing on multimodal transport networks with R5 in R'
author: "Rafael H. M. Pereira, Marcus Saraiva, Daniel Herszenhut, Carlos Kaue Braga"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
  html_document: default
abstract: Routing is a key step in transport planning. Nonetheless, researchers and
  practitioners still face challenges to perform this task due the costs of licensed
  software or the long computation times to run multiple routing analysis in complex
  transport networks. Conveyal's R<sup>5</sup> is a multimodal transport network router
  that offers meaningful planning features, such as accounting for service reliability
  issues and returning multiple travel time estimates and itineraries for a single
  or multiple origin/destination pairs. This paper describes r5r, an open source R
  package that leverages on R<sup>5</sup> to allow users to efficiently compute travel
  time matrices and generate detailed itineraries between a set of origins and destinations
  at no expense. The development of this package contributes to the transport field
  as by offerin an easy-to-use powerful tool for robust planning and research.
urlcolor: blue
bibliography: references.json
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, 
  fig.height = 5
)

set.seed(0)
```

<!-- Examples of similar papers -->

<!--  - https://transportfindings.org/article/8416-accessibility-toolbox-for-r-and-arcgis -->

<!--  - https://transportfindings.org/article/6945-dodgr-an-r-package-for-network-flow-aggregation -->

<!--  - https://joss.theoj.org/papers/10.21105/joss.01926 -->

# RESEARCH QUESTIONS AND HYPOTHESES

Transport routing is the process of finding alternative routes and costs that connect places on a given transport network. Routing often involves complex mathematical algorithms and it is a fundamental step required for transport accessibility analysis, fleet allocation and transport simulation and planning more broadly. Despite the long-standing importance of routing analysis to the field of transport and mobility studies, researchers and practitioners still frequently face practical challenges when carrying out this task. These challenges involve, for example, the monetary costs of licensed software, limited data availability for data hungry models, and the long computation times required to run multiple routing scenarios, particularly in large and complex multimodal transport networks.

While there is a growing number of open source routing models [@opentripplanneropentripplanner; @lovelace2019stplanr; @padgham2019dodgr], most options available do not efficiently process large transport networks. More importantly, those tools generate deterministic estimates of travel times without accounting for variability in results (due to service reliability issues, for example) and only return the fastest route option between a set of origins and destinations (see [@conway2018accounting] for a more in-depth discussion). This simple approach overlooks how often people's actual route choices deviate from shortest travel time paths, what can occur due to multiple factors such as imperfect information, monetary costs and safety issues [@larsen2011travel; @zhu2015people; @conway2019getting].

This paper presents [r5r](https://ipeagit.github.io/r5r/), a new open source R package for rapid realistic routing on multimodal transport networks based on the [Rapid Realistic Routing on Real-world and Reimagined networks (R](https://github.com/conveyal/r5)<sup>5</sup>). R<sup>5</sup> is a powerful next-generation routing engine written in Java and developed at Conveyal [@conway2017evidencebased; @conway2018accounting] to provide an efficient backend for analytic applications - such as accessibility assessment The r5r package provides a simple and friendly interface to run R<sup>5</sup> locally from within R, which allows users to efficiently calculate travel time matrices or generate multiple route alternatives between origins and destinations using seamless parallel computing.

# METHODS AND DATA

The r5r package has low data requirements, enabling routing analysis on different contexts wherever data on street networks from [OpenStreetMap](https://www.openstreetmap.org/) (OSM) is available. The package is also easily scalable allowing for fast computation at either the city or country levels of analysis. For multimodal transport routing, r5r will automatically combine OSM data with one or multiple public transport data sets provided by the user in the standard [General Transit Feed Specification](https://developers.google.com/transit/gtfs/) (GTFS) format to create a routable transport network.

The r5r package has 3 fundamental functions:

-   `setup_r5()`: builds a multimodal transport network used for routing in R<sup>5</sup>. This function automatically (1) downloads/updates a compiled R<sup>5</sup> JAR file, tailored for the purposes of the package, and stores it locally in the r5r package directory for future use; and (2) combines the OSM and GTFS data sets to build a routable network object.

-   `travel_time_matrix()`: computes travel time estimates between one or multiple origin/destination pairs for a single departure time or for multiple departure times over a time window. This function uses an R5-specific extension to the RAPTOR routing algorithm which provides an efficient and systematic sample of multiple departures per minute over a `time_window` set by the user [@conway2017evidencebased]. If the user passes a `time_window` parameter to the function, r5r calculates multiple travel time matrices departing each minute and return additional information on the percentile distribution of travel times that allows one to grasp service reliability issues.

-   `detailed_itineraries()`: computes detailed information on either the shortest or multiple alternative routes between one or multiple origin/destination pairs for a single departure time. The output includes detailed information on route alternatives such as the transport mode, waiting time, travel time and distance of each segment of the trip. This function uses an R5-specific extension to the McRAPTOR routing algorithm to find optimal or less than optimal paths. The specific extension to McRAPTOR to do suboptimal path routing are not documented yet, but a detailed description of base McRAPTOR can be found in [@delling2015roundbased].

Both routing functions are versatile so users can easily set customized inputs such as transport modes, departure date and times, walking and cycling speeds, maximum trip duration, walking distances and number of public transport transfers. We will focus on results obtained from `travel_time_matrix()` on the findings discussed in the following section. 

# FINDINGS

The package can be installed with the following command[^1]:

```{r, message = FALSE, eval = FALSE}
# stable version
install.packages("r5r")
```

By default `R` allocates only 512 MB of memory to the Java Virtual Machine (JVM), which might not be enough to perform tasks on large multimodal networks. To increase the amount of memory available to the package, the user needs to set the `java.parameters` option at the beginning of the script. After that, the package can be attached (alongside some other packages used in this article), as follows:

[^1]: To use r5r, one needs to have locally installed Java SE Development Kit, version 11.0.8 or higher. Available at <https://www.oracle.com/java/technologies/javase-jdk11-downloads.html>.

```{r, message = FALSE}
options(java.parameters = "-Xmx2G")

library(r5r)
library(sf)
library(data.table)
library(ggplot2)
```

The package includes sample datasets for the cities of São Paulo and Porto Alegre (both in Brazil), which are used in the package's vignette and examples to illustrate its functionalities. Each dataset includes:

-   An OSM network in `.pbf` format.
-   A public transport network in `GTFS.zip` format.
-   The spatial coordinates of points covering the area in `.csv` format.

### Build a routable transport network

To build a routable transport network with r5r and load it in memory, the user needs one command line pointing to the location on disk where the input data is stored. In the examples of this paper, we will be using the provided Porto Alegre dataset.

```{r, message = FALSE}
# points to directory with data
data_path <- system.file("extdata/poa", package = "r5r")

r5r_core <- setup_r5(data_path, verbose = FALSE)
```

The function uses the `.pbf` and the `GTFS.zip` files in the directory pointed by `data_path` to create a multimodal transport network used for routing by R<sup>5</sup>. Multiple GTFS feeds can be used simultaneously, in which case R<sup>5</sup> will automatically merge all files into a single public transport network. The resulting `network.dat` as well as some other files used by R<sup>5</sup> are saved inside the supplied directory for later reuse.

### Calculate a travel time matrix

Calculating a travel time matrix with r5r can be done with the `travel_time_matrix()` function. This function takes, as inputs, the spatial location of origins/destinations  (either as a spatial `sf POINT` object, or as a `data.frame` containing the columns `id`, `lon` and `lat`) and a few travel parameters such as maximum trip duration or walking distance. 

By default, `r5r` will return a single travel time estimate for each origin-destination pair at a set `departure_datetime`. However, this can provide biased estimates due to how service levels varies across the day (reference to https://www.sciencedirect.com/science/article/pii/S0966692318305209). To account
for these variations, the user can set a `time_window` in the `travel_time_matrix()` function. In this case, several matrices are calculated using multiple departure
times every minute within the set time window. By default the function outputs the 50th percentile (i.e. the median) travel time between each origin and destination pair  within that time window. Nonetheless, it is possible to retrieve information on travel time estimates at different points of the distribution using the `percentiles` parameter. The function outputs a `data.table` with the travel time estimates between all combinations of origins and destinations that can be reached with the given parameters. An example of the `travel_time_matrix()` function's usage is presented below:

```{r, message = FALSE}
# read points of origin and destination
points <- fread(file.path(data_path, "poa_hexgrid.csv"))

mode <- c("WALK", "TRANSIT")
max_walk_dist <- 1000
max_trip_duration <- 120
departure_datetime <- as.POSIXct("13-05-2019 14:00:00",
                                 format = "%d-%m-%Y %H:%M:%S")

time_window <- 120
percentiles <- c(20, 40, 60, 80, 99)

# calculate travel time matrix
ttm <- travel_time_matrix(r5r_core,
                          origins = points,
                          destinations = points,
                          mode = mode,
                          departure_datetime = departure_datetime,
                          max_walk_dist = max_walk_dist,
                          max_trip_duration = max_trip_duration,
                          time_window = time_window,
                          percentiles = percentiles,
                          verbose = FALSE)

head(ttm)
```

#### Accounting for travel-time uncertainty

In the example above we calculated how the estimates of travel-time to arrive at the central bus station vary by place of trip origin. 

We can visualize how the estimated travel times are more uncertain when leaving from some places than others. The plot below shows travel times from all origin points to the central bus station at five percentiles (20th, 40th, 60th, 80th, 99th). As expected, there is little to no uncertainty in travel times from places that are very close (walking distance) to the central bus station. Conversely, travel times from places that are far away are more affected by variations in the trip's departure time and service frequency levels.

```{r, message = FALSE}
# select the travel time matrix departing from a given origin
central_bus_stn <- points[291,]
ttm_tw <- ttm[toId %in% central_bus_stn$id ] 

# reshape data
plot_data <- melt(ttm_tw, measure = patterns("^travel_time_p"), variable = "percentile", value = "travel_time")
plot_data[, mediantt := median(travel_time, na.rm=T), by=fromId]

ggplot(data=plot_data, aes(y = travel_time, x = reorder(fromId, mediantt))) +
  geom_point(alpha = 0.1, size = .7) +
  geom_line(aes(y=mediantt, group=toId), color="#FE9F45", size=1.5) +
  expand_limits(y = 120) +
  scale_y_continuous(breaks = c(0, 30, 60, 90, 120)) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.title.x = element_blank(), 
        axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.border = element_rect(fill = NA, colour = "grey80", size=1)) +
  labs(title = "Travel time uncertainty per destination",
       y = "Travel Time (min)", x='Destinations')
```

#### Visualize Isochrones

If the destination points cover the entire study area, the output can easily be used to build ishochrone-like visualizations. In our example, we can visualize the isochrone departing from the central bus station as follows:

```{r}
cutoffs <- seq(0, 120, 20)

# select the travel time matrix departing from a given origin
central_bus_stn <- points[291,]
ttm_iso <- ttm[fromId %in% central_bus_stn$id ] 

# aggregate travel-times
ttm_iso[, isochrones := cut(x=travel_time_p099, breaks=cutoffs), by= fromId]
  
# join travel-matrix results to points
ttm_iso[points, on=c('toId' ='id'), `:=`(lon = i.lon, lat = i.lat)]

ggplot(ttm_iso, aes(x=lon, y=lat, colour = isochrones)) +
   geom_point(size = 2) +
   geom_point(data = central_bus_stn, aes(x=lon, y=lat), colour="blue", size = 2) +
   coord_map() +
   scale_colour_brewer(palette = "Reds") +
   facet_wrap(~fromId) +
   theme_minimal() +
   theme(axis.title = element_blank(),
         panel.border = element_rect(fill = NA, colour = "grey40"))
```


# Acknowledgments

The [R](https://github.com/conveyal/r5)<sup>5</sup> routing engine is developed at [Conveyal](https://www.conveyal.com/) with contributions from several developers. This work was supported by the Brazilian Institute for Applied Economic Research (Ipea).

# References

<!-- Rafa's notes -->

<!-- jean: #>-  mass of assumptions regarding decision-making and routing (Miller, 2018).  -->

<!-- Haugen, 2011; Schwanen, 2008 -->

<!-- Miller, E.J., 2018. Accessibility: measurement and application in transportation planning. Transp. Rev. 38, #>551–555. https://doi.org/10.1080/01441647.2018.1492778 -->

<!-- This article presents **r5r**, a new open-source `R` package for rapid  -->

<!-- realistic routing on multimodal transport networks. -->
