---
title: 'r5r: Rapid Realistic Routing with R5 in R'
author: "   "
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette: default
  github_document: default
abstract: " write abstract"
urlcolor: blue
vignette: |
  %\VignetteIndexEntry{r5r-intro} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

#> Examples of similar papers
#> - https://transportfindings.org/article/8416-accessibility-toolbox-for-r-and-arcgis
#> 
#> - https://transportfindings.org/article/6945-dodgr-an-r-package-for-network-flow-aggregation
#> 
#> - https://joss.theoj.org/papers/10.21105/joss.01926


# RESEARCH QUESTIONS AND HYPOTHESES 


Transport routing is the process of finding the alternative routes and costs 
that connect places on a given transport network. Routing often involves complex
mathematical algorithms and it is a fundamental step required for transport
accessibility analysis, fleet allocation and transport simulation and planning
more broadly.

Despite the long standing importance of routing analysis to the field, transport 
researchers and practitioners still frequently face practical challenges with 
this task. These challenges involve for example the monetary costs of licensed 
software, data availability issues for data hungry models, and the long 
computation times to run multiple routing scenarios, particularly in large and 
complex multimodal transport networks. While there is a growing number of open source routing models (ref: stplnr, otp, dodgr), most options available only return the fastest route, what overlooks....

...


# METHODS AND DATA


The **r5r** package provides a simple and convenient friendly interface to R<sup>5</sup>, [Rapid Realistic Routing on Real-world and Reimagined networks](https://github.com/conveyal/r5). The  developed by [Conveyal](https://www.conveyal.com/).

papers under R5 we should cite
- https://repository.asu.edu/items/54162
- https://www.jtlu.org/index.php/jtlu/article/view/1074


# FINDINGS

Section to demonstrate how to use the core functions of the package. We can use 
our intro vignette as a starting point and build from there with more the analysis
of larger data sets, include couple more data visualization of results and emphasize
the results of multiple route alternatives.


To illustrate functionality, the package includes a small sample data for the 
city of Porto Alegre (Brazil). It includes three files:
- An Open Street Map network in `.pbf` format (*mandatory*)
- A public transport network in `GTFS.zip` format
- The spatial coordinates of the origin destination pairs in a `.csv` file.



```{r, message = FALSE}
# Load packages
library(r5r)
library(sf)
library(data.table)
library(dplyr)
library(ggplot2)

```


```{r}
path <- system.file("extdata", package = "r5r")
list.files(path)

points <- fread(system.file("extdata/poa_hexgrid.csv", package = "r5r"))
points <- points[ c(sample(1:nrow(points), 10, replace=TRUE)), ]
head(points)

```

By default, R allocates only 512MB of memory for Java processes, which is not enough for large queries using r5r. To increase available memory to 4GB, for example, we need to set the java.parameters option at the beginning of the script, as follows:

```{r, message = FALSE}
options(java.parameters = "-Xmx4G")
```



### Build routable transport network

The first step is to build the multimodal transport used for routing in R5. This
is done with the `setup_r5` function. It does two things: (1) downloads a compiled JAR file
of R5 and store it locally in the r5r package directory; (2) combines the osm.pbf and gtfs.zip data sets to build a routable network object. The function 

```{r, message = FALSE}
# Indicate the path where OSM and GTFS data are stored
r5r_core <- setup_r5(data_path = path)

```

**r5** has two core routing functions.

- `detailed_itineraries`
   * Returns a `data.frame sf LINESTRINGs` with one or multiple alternative routes
   between one or multiple origin destination pairs. The data output brings 
   detailed information on transport mode, travel time, walk distance etc for 
   each trip section
 
- `travel_time_matrix`
   * Fast function that returns a simple 'data.frame' with travel time 
   estimates between one or multiple origin destination pairs.
   
   
## Fast many to many travel time matrix

```{r, message = FALSE}
# calculate a travel time matrix
ttm <- travel_time_matrix( r5r_core = r5r_core,
                           origins = points,
                           destinations = points,
                           trip_date = "2019-05-20",
                           departure_time = "14:00:00",
                           mode = c("WALK", "TRANSIT"),
                           max_street_time = 7200,
                           max_trip_duration = 7200
                          )
head(ttm)
```

## Detailed itineraries

Allows one to get the fastest or multiple alternative routes between a set of
origins and destinations.

Create simple example here and plot the result below

### visualize results


```{r, message = FALSE}
# extract OSM network
street_net <- street_network_to_sf(r5r_core)

# plot
ggplot() +
        geom_sf(data = street_net$edges, color='gray85') +
        theme_minimal()

```












# Acknowledgments

Conveyal for R5


# References




#> Rafa's notes
#> jean: #>-  mass of assumptions regarding decision-making and routing (Miller, 2018). 
#>Haugen, 2011; Schwanen, 2008
#>
#>Miller, E.J., 2018. Accessibility: measurement and application in transportation planning. Transp. Rev. 38, #>551–555. https://doi.org/10.1080/01441647.2018.1492778
#>
#>This article presents **r5r**, a new open-source `R` package for rapid 
#>realistic routing on multimodal transport networks.
